---
title: "microeuk-deepbiosphere-survey"
author: "Sarah Hu"
date: "2/10/2022"
output: 
  html_document:
    number_sections: true
    theme: spacelab
    highlight: monochrome
    collapsed: false
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: true
    toc_depth: '3'
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = FALSE, message = FALSE)
# cache = TRUE
```

# Background
Preliminary analyses for microeukaryote tag-sequence survey.

*Description*: Investigate the diversity of single-celled microbial eukaryotic communities across several deep-sea hydrothermal vent sites (including NE Pacific, Caribbean). We plan to address questions related to the environmental factors that shape protistan community dynamics, and determine if patterns in species diversity and distribution vary at different deep-sea habitats. These questions will be addressed using similarly generated metabarcoding data from several distinct hydrothermal vents. Along with characterizing community structure, we plan to evaluate interactions between protist species (to identify putative predator-prey or parasite-host relationships) and their environment (to explore their relationship to geochemical properties).

*Questions to address*

> What is the general biogeography and distribution of the deep-sea hydrothermal vent microbial eukaryotic community?

> What community structure features (i.e., species richness, proportion cosmopolitan versus endemic, species evenness) are shared across or unique to deep-sea hydrothermal vent sites?

> What environmental features (i.e., temperature, geochemistry) influence microbial eukaryotic community diversity? Can we identify if certain environmental factors select for putative vent endemics?


## Set up R environment

All upstream processing of sequences from QIIME2 can be found [here](https://github.com/shu251/microeuk-amplicon-survey/tree/main/seq-analysis).

```{r Library loading initial, echo=FALSE, message=FALSE}
library(tidyverse); library(cowplot)
library(phyloseq); library(decontam)
library(compositions); library(patchwork); library(ggpubr)
library(ggupset); library(gt); library(treemapify)
library(plotly); library(viridis); library(vegan)
```

```{r set site variables}
axial <- c("Axial")
mcr <- c("VonDamm", "Piccard")
gr <- c("GordaRidge")
all <- c("Axial", "VonDamm", "Piccard", "GordaRidge")
load("data-input/asv-tables-processed-27012022.RData", verbose = TRUE)
metadata <- read.delim("data-input/samplelist-metadata.txt", na.strings = "")
```

# Overall community composition

Investigate overall protistan community composition.

## Relative abundance bar plot

First, establish order and color key for all taxonomic visuals.
```{r}
all_taxa_order <- c("Alveolata-Ciliophora", "Alveolata-Dinoflagellata", "Protalveolata", "Other Alveolata", "Amoebozoa", "Apusozoa", "Excavata", "Hacrobia", "Archaeplastida", "Rhizaria", "Rhizaria-Radiolaria", "Rhizaria-Cercozoa", "Stramenopiles", "Stramenopiles-Opalozoa", "Stramenopiles-Sagenista", "Stramenopiles-Ochrophyta", "Opisthokonta", "Unknown Eukaryota")
# length(all_taxa_order)

all_taxa_color = c("#fa9fb5", "#c51b8a", "#67000d", "#ef3b2c", "#ffffcc", "#feb24c","#cc4c02", "#c7e9b4", "#238b45", "#c6dbef", "#99d8c9","#1d91c0", "#253494", "#9e9ac8", "#54278f", "#bdbdbd", "#252525", "#f0f0f0")
# length(all_taxa_color)
```


```{r Barplot function, tidy=FALSE, fig.height=12, fig.width=9}
alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")
bkgd <- c("Deep seawater", "BSW", "Shallow seawater", "Near vent BW")
plume <- c("Candelabra Plume", "Mt Edwards Plume", "Plume", "Anemone Plume")

all_taxa_color_protalve = c("#fa9fb5", "#c51b8a","#f0f0f0", "#67000d", "#ef3b2c", "#ffffcc", "#feb24c","#cc4c02", "#c7e9b4", "#238b45", "#c6dbef", "#99d8c9","#1d91c0", "#253494", "#9e9ac8", "#54278f", "#bdbdbd", "#252525", "#f0f0f0")

make_bar_relabun <- function(df, selection){
  df_out <- df %>%
  filter(SITE %in% selection) %>% 
  filter(Domain == "Eukaryota") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  group_by(FeatureID, Taxon, Domain, Supergroup, Phylum, Class, Order, Family, Genus, Species,
           VENT, SITE, SAMPLETYPE, YEAR, DATASET, SUPERGROUP) %>% 
    summarise(SEQ_AVG_REP = mean(value)) %>% 
  ungroup()
df_out %>% 
  group_by(SITE, SAMPLETYPE, VENT, YEAR, SUPERGROUP) %>% 
    summarise(SEQ_SUM = sum(SEQ_AVG_REP)) %>% 
  mutate(SUPERGROUP_ORDER = factor(SUPERGROUP, levels = all_taxa_order)) %>% 
  ggplot(aes(x = VENT, y = SEQ_SUM, fill = SUPERGROUP_ORDER)) +
    geom_bar(stat = "identity", position = "fill", color = "black", width = 0.9) +
    facet_grid(. ~ SITE +YEAR + SAMPLETYPE, scale = "free", space = "free") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold", color = "black"),
        strip.background = element_blank(), 
        strip.text = element_text(color = "black", face = "bold", angle = 90),
        axis.title = element_text(color = "black", face = "bold"),
        axis.text.y = element_text(color = "black", face = "bold"),
        legend.position = "right",
        legend.title = element_blank()) +
  scale_y_continuous(expand = c(0,0)) +
    scale_fill_manual(values = all_taxa_color_protalve) +
  labs(x = "", y = "Relative sequence abundance")
}
```


```{r Barplot all, tidy=FALSE, fig.height=8, fig.width=14}
# svg("../../../Manuscripts_presentations_reviews/deepsea-microeuk-ME/vectorgraphic-figures/figs2-barplotall.svg", h = 9, w = 13)
all <- c("Axial", "VonDamm", "Piccard", "GordaRidge")
make_bar_relabun(insitu_asv_wClass, all)
# dev.off()
```

## Heat map at class level
```{r, fig.height=21, fig.width=15}
alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")
bkgd <- c("Deep seawater", "BSW", "Shallow seawater", "Near vent BW")
plume <- c("Candelabra Plume", "Mt Edwards Plume", "Plume", "Anemone Plume")

tmp_phyla <- insitu_asv_wClass %>%
  filter(Domain == "Eukaryota") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  mutate(SUPERGROUP_ORDER = factor(SUPERGROUP, levels = all_taxa_order)) %>% 
  group_by(FeatureID, Taxon, Domain, Supergroup, Phylum, Class,
           SITE, SAMPLETYPE, VENT, YEAR, SUPERGROUP_ORDER) %>% 
    summarise(SEQ_AVG_REP = mean(value)) %>% 
  ungroup() %>% 
  group_by(SITE, SAMPLETYPE, VENT, YEAR) %>% 
    mutate(TOTAL = sum(SEQ_AVG_REP)) %>% 
  ungroup(SITE, SAMPLETYPE, VENT, YEAR) %>% 
  group_by(SITE, SAMPLETYPE, VENT, YEAR, SUPERGROUP_ORDER, Phylum, Class) %>% 
    summarise(SEQ_SUM = sum(SEQ_AVG_REP),
              REL_ABUN = (SEQ_SUM/TOTAL)) %>% 
  distinct()

###
# ggplot(tmp_phyla, aes(x = VENT, y = Class, fill = (REL_ABUN))) +
#     geom_tile(stat = "identity", color = "black") +
#     facet_grid(SUPERGROUP_ORDER + Phylum ~ SITE +YEAR + SAMPLETYPE, scale = "free", space = "free") +
#   theme_linedraw() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
#         strip.background = element_blank(), strip.text = element_text(color = "black"),
#         legend.position = "right",
#         strip.text.y = element_text(angle = 0)) +
#   scale_fill_gradient(low = "white", high = "red") +
#   labs(x = "", y = "Relative abundance")
```


### Stats on sequences and ASVs
Out of all the ASVs and sequences, what percentage of ASVs were found at all 4 of my vent sites? what % of ASVs were unique to individual sites? what % of sequences?
```{r}
totalasvs <- length(unique(insitu_asv_wClass$FeatureID)); totalasvs
totalseq <- sum(insitu_asv_wClass$value); totalseq

unique(insitu_asv_wClass$SITE_CLASS)
tmp_unique <- filter(insitu_asv_wClass, grepl(" only", SITE_CLASS))
tmp_shared <- filter(insitu_asv_wClass, !(grepl(" only", SITE_CLASS)))
tmp_allshared <- filter(insitu_asv_wClass, SITE_CLASS == "All sites")

# Total number of ASVs that were unique to an individual vent field, includes "only" in the site classification.
a <- length(unique(tmp_unique$FeatureID)); a
100*(a/totalasvs)
# % of sequences
a_sum <- sum(tmp_unique$value); a_sum
100*(a_sum/totalseq)


# Total number of ASVs that were found in more than 1 sample
b <- length(unique(tmp_shared$FeatureID)); b
100*(b/totalasvs)
b_sum <- sum(tmp_shared$value); b_sum
100*(b_sum/totalseq)

# Total number of ASVs  at all sites
c <- length(unique(tmp_allshared$FeatureID)); c
100*(c/totalasvs)
c_sum <- sum(tmp_allshared$value); c_sum
100*(c_sum/totalseq)


# Select ASVs found in 1 field and at vent only
tmp_1field_vent <- filter(insitu_asv_wClass, grepl(" only", SITE_CLASS) & CLASS == "Vent only")
d <- length(unique(tmp_1field_vent$FeatureID)); d
100*(d/totalasvs)
d_sum <- sum(tmp_1field_vent$value); d_sum
100*(d_sum/totalseq)
```

## PCA analysis with all samples
```{r, fig.height=5, fig.width=6}
df_wide_asv <- insitu_asv_wClass %>%
  filter(SITE %in% all) %>%
  filter(Domain == "Eukaryota") %>% 
  # Average across replicates
    group_by(FeatureID, SAMPLENAME, VENT) %>% 
    summarise(AVG = mean(value)) %>% 
    ungroup() %>% 
  separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp"), remove = TRUE) %>% 
  mutate(REGION = case_when(
    SITE == "GordaRidge" ~ "Gorda Ridge",
    SITE %in% mcr ~ "Mid-Cayman Rise",
    SITE == "Axial" ~ "Axial")) %>% 
  mutate(VENTNAME = case_when(
    REGION == "Gorda Ridge" ~ VENT,
    REGION == "Mid-Cayman Rise" ~ paste(SITE, VENT, sep = " "),
    REGION == "Axial" ~ paste(VENT, YEAR, sep = " ")
  )) %>% select(-Sample_tmp) %>% 
  unite(SAMPLE, SAMPLETYPE, SITE, VENTNAME, sep = "_", remove = FALSE) %>% 
   group_by(FeatureID, SAMPLE) %>% 
   summarise(SUM = sum(AVG)) %>% 
  pivot_wider(names_from = FeatureID, values_from = SUM, values_fill = 0) %>% 
  column_to_rownames(var = "SAMPLE")
  # look at eigenvalues
  pca_lr <- prcomp(data.frame(compositions::clr(df_wide_asv)))
  variance_lr <- (pca_lr$sdev^2)/sum(pca_lr$sdev^2)
  ## View bar plot
  barplot(variance_lr, main = "Log-Ratio PCA Screeplot", xlab = "PC Axis", ylab = "% Variance", 
    cex.names = 1.5, cex.axis = 1.5, cex.lab = 1.5, cex.main = 1.5)
  ## Extract PCR points
  data.frame(pca_lr$x, SAMPLE = rownames(pca_lr$x)) %>% 
      separate(SAMPLE, c("SAMPLETYPE", "REGION", "VENTNAME"), sep = "_", remove = FALSE) %>% 
  ## Generate PCA plot
  ggplot(aes(x = PC1, y = PC2, shape = SAMPLETYPE, fill = REGION)) +
    geom_hline(yintercept = 0) + geom_vline(xintercept = 0, color = "#525252") +
    geom_point(size=3, stroke = 1, aes(fill = REGION)) +
    ylab(paste0('PC2 ',round(variance_lr[2]*100,2),'%')) +
    xlab(paste0('PC1 ',round(variance_lr[1]*100,2),'%')) +
    scale_shape_manual(values = c(21, 23, 24)) +
    scale_fill_manual(values = c("#fdbb84", "#31a354", "#ef3b2c", "#02818a")) +
    theme_bw() +
    theme(axis.text = element_text(color="black", size=12),
          legend.title = element_blank(),
          axis.title = element_text(color="black", size=14),
          legend.text = element_text(color = "black", size = 14),
          plot.margin = margin(2, 1, 2, 1, "cm")) +
    guides(fill = guide_legend(override.aes = list(shape = 21) ),
              shape = guide_legend(override.aes = list(fill = "black")))
```

## Dendrogram Jaccard dissimilarity
```{r}
df <- as.data.frame(t(df_wide_asv))
names(df)
###
  colnames(df) <- gsub(x = names(df), pattern = "_", replacement = " ")
  colnames(df) <- gsub(x = names(df), pattern = "Vent Axial", replacement = "Axial")
  colnames(df) <- gsub(x = names(df), pattern = "Vent GordaRidge", replacement = "Gorda Ridge")
  colnames(df) <- gsub(x = names(df), pattern = "GordaRidge", replacement = "Gorda Ridge")
  colnames(df) <- gsub(x = names(df), pattern = "Vent VonDamm VonDamm", replacement = "Von Damm")
  colnames(df) <- gsub(x = names(df), pattern = "Vent Piccard Piccard", replacement = "Piccard")
  colnames(df) <- gsub(x = names(df), pattern = "Background Gorda Ridge", replacement = "Gorda Ridge")
  colnames(df) <- gsub(x = names(df), pattern = "Plume Gorda Ridge", replacement = "Gorda Ridge")
  colnames(df) <- gsub(x = names(df), pattern = "VonDamm VonDamm", replacement = "Von Damm")
  colnames(df) <- gsub(x = names(df), pattern = "Piccard Piccard", replacement = "Piccard")
  colnames(df) <- gsub(x = names(df), pattern = " BSW", replacement = "")
  colnames(df) <- gsub(x = names(df), pattern = "Background Axial Deep seawater 2015", replacement = "Background Axial 2015")
  colnames(df) <- gsub(x = names(df), pattern = "Plume Axial Anemone Plume 2015", replacement = "Axial Anemone Plume 2015")
  colnames(df) <- gsub(x = names(df), pattern = "Rav2", replacement = "Ravelin #2")
  colnames(df) <- gsub(x = names(df), pattern = "Plume Von Damm", replacement = "Von Damm")
  colnames(df) <- gsub(x = names(df), pattern = "Plume Piccard", replacement = "Piccard")
# names(df)

# Write over same data frame - fix sample names
dendro_input <- df
# head(dendro_input)
```

Estimate Jaccard distance
```{r}
dendro_jacc <- vegan::vegdist(t(dendro_input), method = "jaccard")
# dendro_jacc <- vegan::vegdist(t(compositions::clr(dendro_input)), method = "euclidean")

cluster_jacc <- hclust(dist(t(dendro_jacc)), method = "average")

library(ggdendro)
dendro_plot_df <- ggdendro::dendro_data(as.dendrogram(cluster_jacc), type = "rectangle")
label_dendro_order <- as.character(dendro_plot_df$labels$label)
# label_dendro_order
```

Check stress
```{r}
# euc_test <- vegan::vegdist(t(compositions::clr(dendro_input)), method = "euclidean")
# 
# jac_test <- vegan::vegdist(t(dendro_input), method = "jaccard")
# 
# set.seed(2930) # set seed
# ## So we can compare the relative composition based distance metric to the presence/absence based distance metric
# euc_nmds <- metaMDS(euc_test, k=2, autotransform=FALSE)
# jac_nmds <- metaMDS(jac_test, k=2, autotransform=FALSE)
# 
# euc_nmds$stress
# jac_nmds$stress
```


Plot dendrogram
```{r, fig.height=10, fig.width=8}
dendro_plot_output <- ggplot(segment(dendro_plot_df)) + 
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) + 
  coord_flip() + 
  scale_y_reverse(expand = c(0.2, 0.5), breaks = c(0, 0.2, 0.4, 0.6, 0.8)) + 
  geom_text(aes(x = x, y = y, label = label, angle = 0, hjust = 0), data = label(dendro_plot_df)) + 
  theme_dendro() + 
  labs(y = "Dissimilarity", title = "Jaccard distance") + 
  theme(axis.text.x = element_text(color = "black", size = 14), 
        axis.line.x = element_line(color = "#252525"), 
        axis.ticks.x = element_line(), axis.title.x = element_text(color = "black", size = 14))
dendro_plot_output
```

Add bar plot in the same order to show proportion of resident versus cosmpolitan ASVs in each sample.

### Bar plot to pair with dendrogram
```{r}
# head(insitu_asv_wClass)
# unique(insitu_asv_wClass$SITE_CLASS)
# unique(insitu_asv_wClass$CLASS)

dendro_bar <- insitu_asv_wClass %>% 
  filter(SITE %in% all) %>%
  filter(Domain == "Eukaryota") %>% 
  # Average across replicates
    group_by(FeatureID, SAMPLENAME, VENT, SITE_CLASS) %>% 
    summarise(AVG = mean(value)) %>% 
    ungroup() %>% 
  separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp"), remove = TRUE) %>% 
  mutate(REGION= case_when(
    SITE == "GordaRidge" ~ "Gorda Ridge",
    SITE %in% mcr ~ "Mid-Cayman Rise",
    SITE == "Axial" ~ "Axial")) %>% 
  mutate(VENTNAME = case_when(
    REGION == "Gorda Ridge" ~ VENT,
    REGION == "Mid-Cayman Rise" ~ paste(SITE, VENT, sep = " "),
    REGION == "Axial" ~ paste(VENT, YEAR, sep = " ")
  )) %>% select(-Sample_tmp) %>% 
  unite(SAMPLE, SAMPLETYPE, SITE, VENTNAME, sep = "_", remove = FALSE) %>%
  mutate(SITE_CLASS_2 = case_when(
    SITE_CLASS == "Piccard & Von Damm" ~ "Piccard & Von Damm only",
    SITE_CLASS == "Piccard & Axial" ~ "MCR & Axial",
    SITE_CLASS == "Piccard & Gorda Ridge" ~ "MCR & Gorda Ridge",
    SITE_CLASS == "Von Damm & Axial" ~ "MCR & Axial",
    SITE_CLASS == "Von Damm & Gorda Ridge" ~ "MCR & Gorda Ridge",
    SITE_CLASS == "Piccard, Axial, & Gorda Ridge" ~ "MCR, Axial, & Gorda Ridge",
    SITE_CLASS == "Von Damm, Axial, & Gorda Ridge" ~ "MCR, Axial, & Gorda Ridge",
    TRUE ~ SITE_CLASS
  )) %>% 
  group_by(SITE_CLASS_2, SAMPLE) %>% 
  summarise(SEQ_SUM = sum(AVG),
     ASV_COUNT = n()) %>%  
  mutate(SAMPLE = gsub(SAMPLE, pattern = "_", replacement = " ")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Vent Axial", replacement = "Axial")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Vent GordaRidge", replacement = "Gorda Ridge")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "GordaRidge", replacement = "Gorda Ridge")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Vent VonDamm VonDamm", replacement = "Von Damm")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Vent Piccard Piccard", replacement = "Piccard")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Background Gorda Ridge", replacement = "Gorda Ridge")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Plume Gorda Ridge", replacement = "Gorda Ridge")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "VonDamm VonDamm", replacement = "Von Damm")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Piccard Piccard", replacement = "Piccard")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = " BSW", replacement = "")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Background Axial Deep seawater 2015", replacement = "Background Axial 2015")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Plume Axial Anemone Plume 2015", replacement = "Axial Anemone Plume 2015")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Rav2", replacement = "Ravelin #2")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Plume Von Damm", replacement = "Von Damm")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Plume Piccard", replacement = "Piccard"))
  
# unique(insitu_asv_wClass$CLASS)
cosmo <- c("Vent, plume, & background", "Vent & background", "Vent & plume", "Plume & background")

dendro_res_cos_df <- insitu_asv_wClass %>% 
  filter(SITE %in% all) %>%
  filter(Domain == "Eukaryota") %>% 
  mutate(DISTRIBUTION = case_when(
    CLASS %in% cosmo ~ "Cosmopolitan",
    TRUE ~ CLASS
  )) %>% 
  # Average across replicates
    group_by(FeatureID, SAMPLENAME, VENT, DISTRIBUTION) %>% 
    summarise(AVG = mean(value)) %>% 
    ungroup() %>% 
  separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp"), remove = TRUE) %>% 
  mutate(REGION= case_when(
    SITE == "GordaRidge" ~ "Gorda Ridge",
    SITE %in% mcr ~ "Mid-Cayman Rise",
    SITE == "Axial" ~ "Axial")) %>% 
  mutate(VENTNAME = case_when(
    REGION == "Gorda Ridge" ~ VENT,
    REGION == "Mid-Cayman Rise" ~ paste(SITE, VENT, sep = " "),
    REGION == "Axial" ~ paste(VENT, YEAR, sep = " ")
  )) %>% select(-Sample_tmp) %>% 
  unite(SAMPLE, SAMPLETYPE, SITE, VENTNAME, sep = "_", remove = FALSE) %>%
  group_by(DISTRIBUTION, SAMPLE) %>% 
  summarise(SEQ_SUM = sum(AVG),
     ASV_COUNT = n()) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "_", replacement = " ")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Vent Axial", replacement = "Axial")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Vent GordaRidge", replacement = "Gorda Ridge")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "GordaRidge", replacement = "Gorda Ridge")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Vent VonDamm VonDamm", replacement = "Von Damm")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Vent Piccard Piccard", replacement = "Piccard")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Background Gorda Ridge", replacement = "Gorda Ridge")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Plume Gorda Ridge", replacement = "Gorda Ridge")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "VonDamm VonDamm", replacement = "Von Damm")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Piccard Piccard", replacement = "Piccard")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = " BSW", replacement = "")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Background Axial Deep seawater 2015", replacement = "Background Axial 2015")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Plume Axial Anemone Plume 2015", replacement = "Axial Anemone Plume 2015")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Rav2", replacement = "Ravelin #2")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Plume Von Damm", replacement = "Von Damm")) %>% 
  mutate(SAMPLE = gsub(SAMPLE, pattern = "Plume Piccard", replacement = "Piccard"))


dendro_bar$SAMPLE_ORDER <- factor(dendro_bar$SAMPLE, levels = label_dendro_order) 
dendro_res_cos_df$SAMPLE_ORDER <- factor(dendro_res_cos_df$SAMPLE, levels = label_dendro_order)

dendro_bar$SITE_CLASS_ORDER <- factor(dendro_bar$SITE_CLASS_2, levels = c("All sites","Von Damm only","Piccard only","Piccard & Von Damm only","MCR & Axial","MCR & Gorda Ridge", "MCR, Axial, & Gorda Ridge", "Gorda Ridge only","Axial only","Axial & Gorda Ridge"))
```


```{r}
dendro_bar_plot <- ggplot(dendro_bar, aes(x = SAMPLE_ORDER, y = ASV_COUNT, fill = SITE_CLASS_ORDER)) +
  geom_bar(stat = "identity", position = "fill", color = "black", width = 0.7, alpha = 0.8) +
  coord_flip() +
  scale_y_continuous(expand = c(0,0)) +
  # scale_fill_manual(values = c("grey", "#e6550d", "#fdbb84", "#31a354", "#1c9099", "#fde0dd", "#c51b8a")) +
  scale_fill_manual(values = c("#636363","#bd0026","#fed976","#fd8d3c","#a63603","#fdae6b","#c994c7","#00441b","#ce1256","#addd8e","#f7fcb9","#016c59","#41ab5d","#6a51a3","#3690c0")) +
    theme_bw() +
    theme(axis.text = element_text(color="black", size=5),
          legend.title = element_blank(),
          panel.border = element_blank(),
          panel.grid = element_blank(),
          axis.title = element_text(color="black", size=14),
          legend.text = element_text(color = "black", size = 12),
          plot.margin = margin(1, 1, 1, 1, "cm"),
          legend.position = "top") +
  labs(x = "", y = "Proportion of ASVs")

dendro_bar_plot_res_cos <- ggplot(dendro_res_cos_df, aes(x = SAMPLE_ORDER, y = ASV_COUNT, fill = DISTRIBUTION)) +
  geom_bar(stat = "identity", position = "fill", color = "black", width = 0.7, alpha = 0.8) +
  coord_flip() +
  scale_y_continuous(expand = c(0,0)) +
  viridis::scale_fill_viridis(discrete=TRUE, option = "H") +
    theme_bw() +
    theme(axis.text = element_text(color="black", size=5),
          legend.title = element_blank(),
          panel.border = element_blank(),
          panel.grid = element_blank(),
          axis.title = element_text(color="black", size=14),
          legend.text = element_text(color = "black", size = 12),
          plot.margin = margin(1, 1, 1, 1, "cm"),
          legend.position = "top") +
  labs(x = "", y = "Proportion of ASVs")
# dendro_bar_plot_res_cos
# ?scale_fill_viridis

dendro_bar_plot_SEQ <- ggplot(dendro_bar, aes(x = SAMPLE_ORDER, y = SEQ_SUM, fill = SITE_CLASS_ORDER)) +
  geom_bar(stat = "identity", position = "fill", color = "black", width = 0.7, alpha = 0.8) +
  coord_flip() +
  scale_y_continuous(expand = c(0,0)) +
  # scale_fill_manual(values = c("grey", "#e6550d", "#fdbb84", "#31a354", "#1c9099", "#fde0dd", "#c51b8a")) +
  scale_fill_manual(values = c("#636363","#bd0026","#fed976","#fd8d3c","#a63603","#fdae6b","#c994c7","#00441b","#ce1256","#addd8e","#f7fcb9","#016c59","#41ab5d","#6a51a3","#3690c0")) +
    theme_bw() +
    theme(axis.text = element_text(color="black", size=5),
          legend.title = element_blank(),
          panel.border = element_blank(),
          panel.grid = element_blank(),
          axis.title = element_text(color="black", size=14),
          legend.text = element_text(color = "black", size = 12),
          plot.margin = margin(1, 1, 1, 1, "cm"),
          legend.position = "top") +
  labs(x = "", y = "Proportion of sequences")

dendro_bar_plot_res_cos_SEQ <- ggplot(dendro_res_cos_df, aes(x = SAMPLE_ORDER, y = SEQ_SUM, fill = DISTRIBUTION)) +
  geom_bar(stat = "identity", position = "fill", color = "black", width = 0.7, alpha = 0.8) +
  coord_flip() +
  scale_y_continuous(expand = c(0,0)) +
  viridis::scale_fill_viridis(discrete=TRUE, option = "H") +
    theme_bw() +
    theme(axis.text = element_text(color="black", size=5),
          legend.title = element_blank(),
          panel.border = element_blank(),
          panel.grid = element_blank(),
          axis.title = element_text(color="black", size=14),
          legend.text = element_text(color = "black", size = 12),
          plot.margin = margin(1, 1, 1, 1, "cm"),
          legend.position = "t∂p") +
  labs(x = "", y = "Proportion of sequences")
# dendro_bar_plot
```
Combine all plots with patchwork.
```{r Combine plots dendro and bar, fig.width=18, fig.height=12}
# library(patchwork)
# svg("dendrogram_wbarplots.svg", w = 18, h = 12)
dendro_plot_output + dendro_bar_plot + dendro_bar_plot_res_cos + patchwork::plot_layout(nrow = 1, widths = c(1, 0.2, 0.2),
                  heights = c(1.5, 0.2, 0.2)) + plot_annotation(tag_levels = "a")
# dev.off()
```

### ASV richness

ASV richness with customized color schema

```{r, fig.height=4, fig.width=6}
insitu_asv_wClass %>% 
  filter(SITE %in% all) %>%
  filter(Domain == "Eukaryota") %>% 
  # Average across replicates
    group_by(FeatureID, SAMPLENAME, VENT, Supergroup) %>% 
    summarise(AVG = mean(value)) %>% 
  separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp"), remove = TRUE) %>% 
  mutate(REGION = case_when(
    SITE == "GordaRidge" ~ "Gorda Ridge",
    SITE %in% mcr ~ "Mid-Cayman Rise",
    SITE == "Axial" ~ "Axial")) %>% 
  mutate(VENTNAME = case_when(
    REGION == "Gorda Ridge" ~ VENT,
    REGION == "Mid-Cayman Rise" ~ paste(SITE, VENT, sep = " "),
    REGION == "Axial" ~ paste(VENT, YEAR, sep = " ")
  )) %>% select(-Sample_tmp) %>% 
  unite(SAMPLE, REGION, VENTNAME, sep = " ", remove = FALSE) %>% 
  ungroup() %>% 
  group_by(SITE, REGION, SAMPLE, SAMPLETYPE) %>% 
    summarise(NUM_ASV = n()) %>% 
  ggplot(aes(x = SAMPLETYPE, y = NUM_ASV, shape = SAMPLETYPE)) +
  geom_boxplot(aes(group = SAMPLETYPE), alpha = 0.8, width = 0.4) +
  geom_jitter(size=2, width = 0.3, aes(fill = SITE)) +
  scale_shape_manual(values = c(21, 23, 24)) +
    scale_fill_manual(values = c("#fdbb84", "#31a354", "#ef3b2c", "#02818a")) +
  theme_bw() +
  theme(axis.text = element_text(color="black", size=12),
        legend.title = element_blank(),
        axis.title = element_text(color="black", size=14),
        legend.text = element_text(color = "black", size = 14)) +
  guides(fill = guide_legend(override.aes = list(shape = 21) ),
            shape = guide_legend(override.aes = list(fill = NA) ) ) +
  labs(x = "", y = "Total number of ASVs")
```

### ASV richness by sample type & taxa
Figure S4.
```{r, fig.height=8, fig.width=12}
# svg("../../../Manuscripts_presentations_reviews/deepsea-microeuk-ME/vectorgraphic-figures/figs4-ASVrichness-bytax.svg", h = 8, w = 12)
insitu_asv_wClass %>% 
  filter(SITE %in% all) %>%
  filter(Domain == "Eukaryota") %>% 
  filter(Domain == "Eukaryota") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>%
  # Average across replicates
    group_by(FeatureID, SAMPLENAME, VENT, SUPERGROUP) %>% 
    summarise(AVG = mean(value)) %>% 
  separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp"), remove = TRUE) %>% 
  mutate(REGION = case_when(
    SITE == "GordaRidge" ~ "Gorda Ridge",
    SITE %in% mcr ~ "Mid-Cayman Rise",
    SITE == "Axial" ~ "Axial")) %>% 
  mutate(VENTNAME = case_when(
    REGION == "Gorda Ridge" ~ VENT,
    REGION == "Mid-Cayman Rise" ~ paste(SITE, VENT, sep = " "),
    REGION == "Axial" ~ paste(VENT, YEAR, sep = " ")
  )) %>% select(-Sample_tmp) %>% 
  unite(SAMPLE, REGION, VENTNAME, sep = " ", remove = FALSE) %>% 
  ungroup() %>% 
  group_by(SITE, REGION, SAMPLE, SAMPLETYPE, SUPERGROUP) %>% 
    summarise(NUM_ASV = n()) %>% 
  mutate(SUPERGROUP_ORDER = factor(SUPERGROUP, levels = all_taxa_order)) %>%
  ggplot(aes(x = SAMPLETYPE, y = NUM_ASV, shape = SAMPLETYPE)) +
  geom_boxplot(aes(group = SAMPLETYPE), alpha = 0.8, width = 0.4) +
  geom_jitter(size=2, width = 0.3, aes(fill = SUPERGROUP_ORDER)) +
  facet_wrap(.~ SUPERGROUP_ORDER, scales = "free_y") +
  scale_shape_manual(values = c(21, 23, 24)) +
    scale_fill_manual(values = all_taxa_color_protalve) +
  theme_bw() +
  theme(axis.text = element_text(color="black", size=12),
        legend.title = element_blank(),
        axis.title = element_text(color="black", size=14),
        legend.text = element_text(color = "black", size = 14),
        strip.background = element_blank(),
        axis.text.x = element_text(color="black", size=12, angle = 90, hjust = 1, vjust = 0.5)) +
  guides(fill = guide_legend(override.aes = list(shape = 21) ),
            shape = guide_legend(override.aes = list(fill = NA) ) ) +
  labs(x = "", y = "Total number of ASVs")
# dev.off()
```

# ASV distribution by sample/site

## UpsetR by ASV level

```{r, fig.width=10, fig.height=8}
alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")
# svg("../../../Manuscripts_presentations_reviews/deepsea-microeuk-ME/vectorgraphic-figures/fig2b-upsetR-all.svg", h=8, w=10)
insitu_asv_wClass %>% 
  filter(SITE %in% all) %>%
  filter(Domain == "Eukaryota") %>% 
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    Supergroup == "Protalveolata" ~ "Other Alveolata",
    TRUE ~ Supergroup
  )) %>% 
  # Taxa to supergroup
  mutate(SupergroupPhylum = SUPERGROUP) %>%
  # Average across replicates
    group_by(FeatureID, SAMPLENAME, VENT, SupergroupPhylum) %>% 
    summarise(AVG = mean(value)) %>% 
  ungroup() %>% 
  separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp"), remove = TRUE) %>% 
  mutate(REGION = case_when(
    SITE == "GordaRidge" ~ "Gorda Ridge",
    SITE %in% mcr ~ "Mid-Cayman Rise",
    SITE == "Axial" ~ "Axial")) %>% 
  mutate(VENTNAME = case_when(
    REGION == "Gorda Ridge" ~ VENT,
    REGION == "Mid-Cayman Rise" ~ paste(SITE, VENT, sep = " "),
    REGION == "Axial" ~ paste(VENT, YEAR, sep = " ")
  )) %>% select(-Sample_tmp) %>% 
  unite(SAMPLE, SITE, SAMPLETYPE, sep = " ", remove = FALSE) %>% 
  group_by(FeatureID, SupergroupPhylum, SAMPLE) %>% 
    summarise(SUM = sum(AVG)) %>%
  # filter(SUM > 200) %>% 
  ungroup() %>% 
  mutate(SUPERGROUP_ORDER = factor(SupergroupPhylum, levels = all_taxa_order)) %>%
  distinct(FeatureID, SUPERGROUP_ORDER, SUM, SAMPLE, .keep_all = TRUE) %>% 
  group_by(FeatureID, SUPERGROUP_ORDER) %>% 
  summarise(SAMPLE = list(SAMPLE)) %>% 
  ggplot(aes(x = SAMPLE)) +
    geom_bar(color = NA, width = 0.7, aes(fill = SUPERGROUP_ORDER)) +
    scale_x_upset(n_intersections = 25) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = "", y = "Shared ASVs") +
  theme_linedraw() +
  theme(axis.text.y = element_text(color="black", size=14, face = "bold"),
        axis.text.x = element_text(color="black", size=14, face = "bold"),
        axis.title = element_text(color="black", size=14, face = "bold"),
        legend.text = element_text(color = "black", size = 12),
        plot.margin = margin(1, 1, 1, 5, "cm"),
        legend.title = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  scale_fill_manual(values = all_taxa_color)
# dev.off()

```

Observations regarding above plot: - Axial and Gorda Ridge vent sites
have more shared ASVs than any other pairwise comparison. After this,
there were also many ASVs shared throughout MCR (vent, plume, +
background). *May be a reflection of sample size, as MCR had more vent
sites* - a small subset of ASVs were found at all vent sites or all
samples. - ASVs within the vents had much higher unique \# of ASVs (not
shared with another habitat type) than any other sample type/location
(furtherest left bars).

### NE Pacific shared 
```{r}
# head(insitu_asv_wClass)
# unique(insitu_asv_wClass$SITE_CLASS)

pacific_shared <- insitu_asv_wClass %>% 
  filter(SITE_CLASS == "Axial & Gorda Ridge" & CLASS == "Vent only")
# View(pacific_shared %>% filter(Phylum == "Ciliophora"))
# View(pacific_shared %>% filter(Phylum == "Cercozoa"))
# View(pacific_shared %>% filter(Phylum == "Stramenopiles"))

axial_yr <- insitu_asv_wClass %>% 
  filter(CLASS == "Vent only") %>% 
  filter(grepl("Axial", SITE_CLASS)) %>% 
  filter(YEAR < 2017) %>% 
  select(YEAR, Taxon, FeatureID, value) %>% 
  pivot_wider(names_from = YEAR, values_from = value, values_fill = NA, values_fn = sum) %>% 
  drop_na()

```

### MCR shared
```{r}
mcr_shared <- insitu_asv_wClass %>% 
  filter(SITE_CLASS == "Piccard & Von Damm" & CLASS == "Vent only")
length(unique(mcr_shared$FeatureID))
```

### Vents all shared
```{r}
allvents_resident <- insitu_asv_wClass %>% 
  filter(SITE_CLASS == "All sites" & CLASS == "Vent only")
length(unique(allvents_resident$FeatureID))
```

## UpsetR at Genus level
Repeat upsetR plot, but summarize at genus level, rather than "species" or "strain"


```{r, fig.width=15, fig.height=10}
alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")
# svg("upsetR-bysite-sampletype-nov2.svg", h=9, w=15)
genus_upset <- insitu_asv_wClass %>% 
  filter(SITE %in% all) %>%
  filter(Domain == "Eukaryota") %>% 
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  # Taxa to supergroup
  mutate(SupergroupPhylum = SUPERGROUP) %>%
  unite(GENUS, Domain:Genus, sep = ";") %>% 
  # Average across replicates
    group_by(GENUS, SAMPLENAME, VENT, SupergroupPhylum) %>% 
    summarise(AVG = mean(value)) %>% 
  ungroup() %>% 
  separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp"), remove = TRUE) %>% 
  mutate(REGION = case_when(
    SITE == "GordaRidge" ~ "Gorda Ridge",
    SITE %in% mcr ~ "Mid-Cayman Rise",
    SITE == "Axial" ~ "Axial")) %>% 
  mutate(VENTNAME = case_when(
    REGION == "Gorda Ridge" ~ VENT,
    REGION == "Mid-Cayman Rise" ~ paste(SITE, VENT, sep = " "),
    REGION == "Axial" ~ paste(VENT, YEAR, sep = " ")
  )) %>% select(-Sample_tmp) %>% 
  unite(SAMPLE, SITE, SAMPLETYPE, sep = " ", remove = FALSE) %>% 
  group_by(GENUS, SupergroupPhylum, SAMPLE) %>% 
    summarise(SUM = sum(AVG)) %>%
  # filter(SUM > 200) %>% 
  ungroup() %>% 
  distinct(GENUS, SupergroupPhylum, SUM, SAMPLE, .keep_all = TRUE) %>% 
  group_by(GENUS, SupergroupPhylum) %>% 
  summarise(SAMPLE = list(SAMPLE)) %>% 
  ggplot(aes(x = SAMPLE)) +
    geom_bar(color = "black", width = 0.7, aes(fill = SupergroupPhylum)) +
    scale_x_upset(n_intersections = 25) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = "", y = "Shared at Genus level") +
  theme_linedraw() +
  theme(axis.text.y = element_text(color="black", size=14, face = "bold"),
        axis.text.x = element_text(color="black", size=14, face = "bold"),
        axis.title = element_text(color="black", size=14, face = "bold"),
        legend.text = element_text(color = "black", size = 12, face = "bold"),
        legend.title = element_blank(),
        panel.grid.minor = element_blank(),
        plot.margin = margin(1, 1, 1, 5, "cm")) +
  scale_fill_manual(values = all_taxa_color)

#
genus_upset
```

Isolate list of genus level that are shared at all sites.
```{r}
alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")
# svg("upsetR-bysite-sampletype-nov2.svg", h=9, w=15)
shared_genus <- insitu_asv_wClass %>% 
  filter(SITE %in% all) %>%
  filter(Domain == "Eukaryota") %>% 
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  # Taxa to supergroup
  mutate(SupergroupPhylum = SUPERGROUP) %>%
  unite(GENUS, Domain:Genus, sep = ";") %>% 
  # Average across replicates
    group_by(GENUS, SAMPLENAME, VENT, SupergroupPhylum) %>% 
    summarise(AVG = mean(value)) %>% 
  ungroup() %>% 
  separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp"), remove = TRUE) %>% 
  mutate(REGION = case_when(
    SITE == "GordaRidge" ~ "Gorda Ridge",
    SITE %in% mcr ~ "Mid-Cayman Rise",
    SITE == "Axial" ~ "Axial")) %>% 
  mutate(VENTNAME = case_when(
    REGION == "Gorda Ridge" ~ VENT,
    REGION == "Mid-Cayman Rise" ~ paste(SITE, VENT, sep = " "),
    REGION == "Axial" ~ paste(VENT, YEAR, sep = " ")
  )) %>% select(-Sample_tmp) %>% 
  unite(SAMPLE, SITE, SAMPLETYPE, sep = " ", remove = FALSE) %>% 
  select(GENUS, SAMPLE, AVG) %>% 
  pivot_wider(names_from = SAMPLE, values_from = "AVG", values_fn = sum) %>%
  drop_na()

# head(shared_genus)
```

## Resident vs. cosmopolitan stats

### Resident
From the vent-only ASVs, what percentage of them appear only at 1 site? what percent also appear at other vent sites?
```{r}
resident_only <- filter(insitu_asv_wClass, CLASS == "Vent only")

totalasvs <- length(unique(resident_only$FeatureID)); totalasvs
totalseq <- sum(resident_only$value); totalseq

# resident includes
100*(totalasvs/length(unique(insitu_asv_wClass$FeatureID))) #%ASVs
100*(totalseq/sum(insitu_asv_wClass$value)) #%sequences
  # unique(resident_only$SITE_CLASS)
tmp_unique <- filter(resident_only, grepl(" only", SITE_CLASS))
tmp_shared <- filter(resident_only, !(grepl(" only", SITE_CLASS)))
tmp_allshared <- filter(resident_only, SITE_CLASS == "All sites")

# Of the ASVs found ONLY within diffuse venting fluid, just over 90% were unique to only the individual vent site - makes up 53% of the vent only sequences
a <- length(unique(tmp_unique$FeatureID)); a
100*(a/totalasvs)
100*(a/length(unique(insitu_asv_wClass$FeatureID)))

a_sum <- sum(tmp_unique$value); a_sum
100*(a_sum/totalseq)


# Of the ASVs from vent only, 9.6% were also found at other vent sites, totals to 46% of the sequences.
b <- length(unique(tmp_shared$FeatureID)); b
100*(b/totalasvs)

b_sum <- sum(tmp_shared$value); b_sum 
100*(b_sum/totalseq)

# Total number of ASVs designated to be found at all sites
c <- length(unique(tmp_allshared$FeatureID)); c
100*(c/totalasvs)

c_sum <- sum(tmp_allshared$value); c_sum
100*(c_sum/totalseq)

vent_only_allsites <- tmp_allshared
# View(vent_only_allsites)
table(vent_only_allsites$Genus)
unique(vent_only_allsites$Taxon)
# unique(vent_only_allsites$FeatureID)

```


### Cosmopolitan
Isolate ASVs that were found in all habitat types! (Vent, plume, & background)

```{r}
# unique(insitu_asv_wClass$CLASS)
cosmo <- c("Vent, plume, & background", "Vent & background", "Vent & plume", "Plume & background")
# cosmo <- c("Vent, plume, & background")
cosmo_only <- filter(insitu_asv_wClass, CLASS %in% cosmo)

totalasvs <- length(unique(cosmo_only$FeatureID)); totalasvs
totalseq <- sum(cosmo_only$value); totalseq

100*(totalasvs/length(unique(insitu_asv_wClass$FeatureID)))
100*(totalseq/sum(insitu_asv_wClass$value))

# unique(cosmo_only$SITE_CLASS)
tmp_unique <- filter(cosmo_only, grepl(" only", SITE_CLASS))
tmp_shared <- filter(cosmo_only, !(grepl(" only", SITE_CLASS)))
tmp_allshared <- filter(cosmo_only, SITE_CLASS == "All sites")

# Of the xoamopolitan ASVs, the ASVs found unique to a vent field
a <- length(unique(tmp_unique$FeatureID)); a
100*(a/totalasvs)
100*(a/length(unique(insitu_asv_wClass$FeatureID)))

a_sum <- sum(tmp_unique$value); a_sum
100*(a_sum/totalseq)


# Of the ASVs found in all habitat types, 77% of them were also found at another vent site, consisting of 92% of the sequences
b <- length(unique(tmp_shared$FeatureID)); b
100*(b/totalasvs)

b_sum <- sum(tmp_shared$value); b_sum 
100*(b_sum/totalseq)

# Of the ASVS found in all habitat types, 16% of them were found at all vent sites, which made up 40% of the sequences.
c <- length(unique(tmp_allshared$FeatureID)); c
100*(c/totalasvs)

c_sum <- sum(tmp_allshared$value); c_sum
100*(c_sum/totalseq)
```

### Treemap of resident vs. cosmopolitan

```{r, fig.height=6, fig.width=7}
alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")
bkgd <- c("Deep seawater", "BSW", "Shallow seawater", "Near vent BW")
plume <- c("Candelabra Plume", "Mt Edwards Plume", "Plume", "Anemone Plume")

cosmo <- c("Vent, plume, & background", "Vent & background", "Vent & plume", "Plume & background")
res <- c("Vent only")
# head(insitu_asv_wClass)
# unique(insitu_asv_wClass$CLASS)
```

```{r, fig.height=7, fig.width=7}
to_supergroup_CLASS <- insitu_asv_wClass %>% 
  filter(SITE %in% all) %>% 
  mutate(CATEGORY = case_when(
    CLASS %in% cosmo ~ "Cosmopolitan",
    CLASS %in% res ~ "Resident",
    TRUE ~ "Other"
  )) %>% 
  filter(Domain == "Eukaryota") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  mutate(SAMPLETYPEORDER = case_when(
    VENT %in% bkgd ~ "Background",
    VENT %in% plume ~ "Plume",
    TRUE ~ "Vent"
  )) %>% 
  group_by(FeatureID, Taxon, SUPERGROUP,
           CATEGORY) %>% 
    summarise(SEQ_AVG_REP = mean(value)) %>% 
  ungroup() %>% 
  group_by(CATEGORY, SUPERGROUP) %>% 
    summarise(ASV_COUNT = n(),
      SEQ_SUM = sum(SEQ_AVG_REP))

# head(to_supergroup_CLASS)
to_supergroup_CLASS %>% 
  filter(CATEGORY != "Other") %>% 
  pivot_longer(cols = ASV_COUNT:SEQ_SUM) %>% 
  mutate(NAME = case_when(
    name == "SEQ_SUM" ~ "Proportion of sequences",
    TRUE ~ "Proportion of ASVs"
  )) %>% 
  mutate(SUPERGROUP_ORDER = factor(SUPERGROUP, levels = all_taxa_order)) %>%
  ggplot(aes(area = value, fill = SUPERGROUP_ORDER, subgroup = SUPERGROUP_ORDER)) +
    geom_treemap(color = "white") +
    geom_treemap_subgroup_border(colour = "white", size = 2) +
    facet_grid(CATEGORY ~ NAME) +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        strip.background = element_blank(), strip.text = element_text(color = "black"),
        legend.position = "right",
        legend.title = element_blank(),
        panel.border = element_blank()) +
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_manual(values = all_taxa_color) +
  labs(x = "", y = "")

```

### Bubble plot comparison
```{r}
supergroup_distri_bubble <- insitu_asv_wClass %>% 
  filter(SITE %in% all) %>% 
  mutate(CATEGORY = case_when(
    CLASS %in% cosmo ~ "Cosmopolitan",
    CLASS %in% res ~ "Resident",
    TRUE ~ "Other"
  )) %>% 
  filter(Domain == "Eukaryota") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  mutate(SAMPLETYPEORDER = case_when(
    VENT %in% bkgd ~ "Background",
    VENT %in% plume ~ "Plume",
    TRUE ~ "Vent"
  )) %>% 
  group_by(SUPERGROUP, CATEGORY) %>%
    summarise(SEQ_TOTAL = sum(value),
              ASV_TOTAL = n()) %>% 
  filter(SEQ_TOTAL > 200) %>%
  ungroup() %>% 
  pivot_wider(names_from = CATEGORY, values_from = c(SEQ_TOTAL, ASV_TOTAL), values_fill = 0) %>%
  mutate(SUPERGROUP_SEQ_TOTAL = SEQ_TOTAL_Cosmopolitan + SEQ_TOTAL_Other + SEQ_TOTAL_Resident,
         cosmo_seq_perc = 100*(SEQ_TOTAL_Cosmopolitan/SUPERGROUP_SEQ_TOTAL),
         res_seq_perc = 100*(SEQ_TOTAL_Resident/SUPERGROUP_SEQ_TOTAL),
         other_seq_perc = 100*(SEQ_TOTAL_Other/SUPERGROUP_SEQ_TOTAL),
         SUPERGROUP_ASV_TOTAL = ASV_TOTAL_Cosmopolitan + ASV_TOTAL_Other + ASV_TOTAL_Resident,
         cosmo_asv_perc = 100*(ASV_TOTAL_Cosmopolitan/SUPERGROUP_ASV_TOTAL),
         res_asv_perc = 100*(ASV_TOTAL_Resident/SUPERGROUP_ASV_TOTAL),
         other_asv_perc = 100*(ASV_TOTAL_Other/SUPERGROUP_ASV_TOTAL))
```


```{r, fig.height=4, fig.width=4}
# svg("../../../Manuscripts_presentations_reviews/deepsea-microeuk-ME/vectorgraphic-figures/fig5a-bubble-wlegend.svg", h = 4, w = 4)
supergroup_distri_bubble %>% 
  mutate(SUPERGROUP_ORDER = factor(SUPERGROUP, levels = all_taxa_order)) %>%
  ggplot(aes(y = res_asv_perc, x = cosmo_asv_perc, size = SUPERGROUP_SEQ_TOTAL, fill = SUPERGROUP_ORDER)) +
  # geom_abline(slope = 1, intercept = 0, alpha = 0.5) +
  geom_point(shape = 21, stat = "identity", color = "#525252", stroke = 1) +
  scale_size_continuous(range = c(5,25)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_x_continuous(limits = c(0,100)) +
  theme_linedraw() +
  scale_fill_manual(values = all_taxa_color) +
  labs(x = "Proportion of cosmopolitan ASVs", y = "Proportion of resident ASVs") +
  theme(legend.title = element_blank(),
        # legend.position = "none",
        legend.position = "right", # color key is the same
        axis.text = element_text(color = "black", face = "bold", size = 12),
        axis.title = element_text(color = "black", face = "bold", size = 12),
        panel.background = element_blank())
dev.off()

# View(select(supergroup_distri_bubble, SUPERGROUP, SUPERGROUP_SEQ_TOTAL)) # check legend using these values
```

```{r}
supergroup_distri_bubble %>% 
  mutate(SUPERGROUP_ORDER = factor(SUPERGROUP, levels = all_taxa_order)) %>% 
  ggplot(aes(x = ASV_TOTAL_Cosmopolitan, y = ASV_TOTAL_Resident, size = SUPERGROUP_SEQ_TOTAL, fill = SUPERGROUP_ORDER)) +
  # geom_abline(slope = 1, intercept = 0, alpha = 0.5) +
  geom_point(shape = 21, stat = "identity", color = "black") +
  scale_size_continuous(range = c(4,20)) +
  scale_y_continuous(limits = c(0,4500)) +
  scale_x_continuous(limits = c(0,4500)) +
  theme_linedraw() +
  scale_fill_manual(values = all_taxa_color) +
  labs(x = "Total number of cosmopolitan ASVs", y = "Total number of resident ASVs") +
  theme(legend.title = element_blank())
```


### Bar plot res vs. cosmopolitan

```{r}
seq_res_cosmo_bar <- insitu_asv_wClass %>% 
  # filter(SITE %in% all) %>% 
  mutate(CATEGORY = case_when(
    CLASS %in% cosmo ~ "Cosmopolitan",
    CLASS %in% res ~ "Resident",
    TRUE ~ "Other"
  )) %>% 
  filter(Domain == "Eukaryota") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  mutate(SAMPLETYPEORDER = case_when(
    VENT %in% bkgd ~ "Background",
    VENT %in% plume ~ "Plume",
    TRUE ~ "Vent"
  )) %>% 
  group_by(SUPERGROUP, CATEGORY, SITE) %>%
    summarise(SEQ_TOTAL = sum(value),
              ASV_TOTAL = n()) %>% 
  filter(CATEGORY != "Other") %>% 
  mutate(SUPERGROUP_ORDER = factor(SUPERGROUP, levels = all_taxa_order)) %>%
  ggplot(aes(x = SITE, y = SEQ_TOTAL, fill = SUPERGROUP_ORDER)) +
    geom_bar(stat = "identity", position = "stack", width = 0.4) +
    facet_grid(. ~ CATEGORY) +
  theme_linedraw() +
  scale_fill_manual(values = all_taxa_color_protalve) +
  labs(x = "", y = "Total sequences") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        strip.background = element_blank(),
        strip.text = element_text(color = "black"))
  
```

```{r, fig.width=8, fig.height=5}
insitu_asv_wClass %>% 
  mutate(CATEGORY = case_when(
    CLASS %in% cosmo ~ "Cosmopolitan",
    CLASS %in% res ~ "Resident",
    TRUE ~ "Other"
  )) %>% 
  filter(Domain == "Eukaryota") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  mutate(SAMPLETYPEORDER = case_when(
    VENT %in% bkgd ~ "Background",
    VENT %in% plume ~ "Plume",
    TRUE ~ "Vent"
  )) %>% 
  group_by(SUPERGROUP, CATEGORY, SITE) %>%
    summarise(SEQ_TOTAL = sum(value),
              ASV_TOTAL = n()) %>% 
  filter(CATEGORY != "Other") %>% 
  mutate(SUPERGROUP_ORDER = factor(SUPERGROUP, levels = all_taxa_order)) %>%
  ggplot(aes(x = SITE, y = SEQ_TOTAL, fill = SUPERGROUP_ORDER)) +
    geom_bar(stat = "identity", position = "fill", width = 0.4) +
    facet_grid(. ~ CATEGORY) +
  theme_linedraw() +
  scale_fill_manual(values = all_taxa_color_protalve) +
  labs(x = "", y = "Sequence relative abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        strip.background = element_blank(),
        strip.text = element_text(color = "black"),
        legend.title = element_blank())
  
```

```{r}
asv_res_cosmo_bar <- insitu_asv_wClass %>% 
  # filter(SITE %in% all) %>% 
  mutate(CATEGORY = case_when(
    CLASS %in% cosmo ~ "Cosmopolitan",
    CLASS %in% res ~ "Resident",
    TRUE ~ "Other"
  )) %>% 
  filter(Domain == "Eukaryota") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  mutate(SAMPLETYPEORDER = case_when(
    VENT %in% bkgd ~ "Background",
    VENT %in% plume ~ "Plume",
    TRUE ~ "Vent"
  )) %>% 
  group_by(SUPERGROUP, CATEGORY, SITE) %>%
    summarise(SEQ_TOTAL = sum(value),
              ASV_TOTAL = n()) %>% 
  filter(CATEGORY != "Other") %>% 
  mutate(SUPERGROUP_ORDER = factor(SUPERGROUP, levels = all_taxa_order)) %>%  
  ggplot(aes(x = SITE, y = ASV_TOTAL, fill = SUPERGROUP_ORDER)) +
    geom_bar(stat = "identity", position = "stack", width = 0.4) +
    facet_grid(. ~ CATEGORY) +
  theme_linedraw() +
  scale_fill_manual(values = all_taxa_color_protalve) +
  labs(x = "", y = "Total ASVs") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        strip.background = element_blank(),
        strip.text = element_text(color = "black"))
  
```

Figure S5
```{r, fig.width=9, fig.height=5}
# svg("../../../Manuscripts_presentations_reviews/deepsea-microeuk-ME/vectorgraphic-figures/figs5-res_cosmo_barplot.svg", w = 9, h = 5)
(seq_res_cosmo_bar + theme(legend.position = "none")) + (asv_res_cosmo_bar + theme(legend.position = "none")) + patchwork::plot_layout(nrow = 1) + plot_annotation(tag_levels = "a")
# dev.off()
```



```{r}
vent_only_resident <- c("Excavata", "Apusozoa", "Amoebozoa")
insitu_asv_wClass %>% 
  # filter(SITE %in% all) %>% 
  mutate(CATEGORY = case_when(
    CLASS %in% cosmo ~ "Cosmopolitan",
    CLASS %in% res ~ "Resident",
    TRUE ~ "Other"
  )) %>% 
  filter(Domain == "Eukaryota") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  mutate(SAMPLETYPEORDER = case_when(
    VENT %in% bkgd ~ "Background",
    VENT %in% plume ~ "Plume",
    TRUE ~ "Vent"
  )) %>% 
  filter(SUPERGROUP %in% vent_only_resident) %>% 
  group_by(SUPERGROUP, CATEGORY, SITE) %>%
    summarise(SEQ_TOTAL = sum(value),
              ASV_TOTAL = n()) %>% 
  filter(CATEGORY != "Other") %>% 
  mutate(SUPERGROUP_ORDER = factor(SUPERGROUP, levels = all_taxa_order)) %>%
  ggplot(aes(x = SITE, y = SEQ_TOTAL, fill = SUPERGROUP_ORDER)) +
    geom_bar(stat = "identity", position = "stack", width = 0.4) +
    facet_grid(. ~ CATEGORY) +
  theme_linedraw() +
  scale_fill_manual(values = all_taxa_color) +
  labs(x = "", y = "Total sequences") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        strip.background = element_blank(),
        strip.text = element_text(color = "black"))
  
```
```{r}
vent_only_resident <- c("Excavata", "Apusozoa", "Amoebozoa")
resident_subset <- insitu_asv_wClass %>% 
  # filter(SITE %in% all) %>% 
  mutate(CATEGORY = case_when(
    CLASS %in% cosmo ~ "Cosmopolitan",
    CLASS %in% res ~ "Resident",
    TRUE ~ "Other"
  )) %>% 
  filter(Domain == "Eukaryota") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  mutate(SAMPLETYPEORDER = case_when(
    VENT %in% bkgd ~ "Background",
    VENT %in% plume ~ "Plume",
    TRUE ~ "Vent"
  )) %>% 
  filter(SUPERGROUP %in% vent_only_resident) %>% 
  filter(CLASS == "Vent only")

dim(resident_subset)
table(resident_subset$Class)
```


## ASV prevalence by sequence and taxa
```{r Create taxonomy key for ASV data}
# Not sure if I need this
tax_key <- insitu_asv_wClass %>% 
  select(FeatureID, Taxon, Domain, Supergroup, Phylum, Class, Order, Family, Genus, Species, CLASS, SITE_CLASS) %>% 
  distinct()
```

```{r}
# head(insitu_asv_wClass)
alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")

tmp <- insitu_asv_wClass %>% 
  filter(Domain == "Eukaryota") %>% 
  filter(!is.na(Supergroup)) %>% 
  # Average across replicates
    group_by(FeatureID, SAMPLENAME, VENT) %>% 
    summarise(AVG = mean(value)) %>% 
  ungroup() %>% 
  separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp"), remove = TRUE) %>% 
  mutate(REGION = case_when(
    SITE == "GordaRidge" ~ "Gorda Ridge",
    SITE %in% mcr ~ "Mid-Cayman Rise",
    SITE == "Axial" ~ "Axial")) %>% 
  mutate(VENTNAME = case_when(
    REGION == "Gorda Ridge" ~ VENT,
    REGION == "Mid-Cayman Rise" ~ paste(SITE, VENT, sep = " "),
    REGION == "Axial" ~ paste(VENT, YEAR, sep = " ")
  )) %>% select(-Sample_tmp) %>% 
  unite(SAMPLE, REGION, VENTNAME, sep = " ", remove = FALSE) %>% 
  group_by(FeatureID, SAMPLE) %>% 
  summarise(SUM = sum(AVG)) %>% 
  pivot_wider(names_from = "SAMPLE", values_from = SUM, values_fill = 0) %>% 
  column_to_rownames(var = "FeatureID") %>%
  mutate(PREVALENCE = rowSums(. > 0),
         SEQ_TOTAL = rowSums(.)) %>% 
  rownames_to_column(var = "FeatureID") %>% 
  left_join(tax_key) %>% 
   filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  mutate(SUPERGROUP_ORDER = factor(SUPERGROUP, levels = all_taxa_order))
```

```{r, fig.height=7, fig.width=9}
# head(tmp)
tmp %>% 
  filter(SEQ_TOTAL > 0) %>% 
  ggplot(aes(x = PREVALENCE, y = SEQ_TOTAL, fill = SUPERGROUP_ORDER)) +
  geom_jitter(stat = "identity", shape = 21) +
  scale_y_log10() +
  facet_wrap(SUPERGROUP_ORDER ~ .) +
  scale_fill_manual(values = all_taxa_color) +
  theme_linedraw() +
  labs(x = "Number of samples ASV appears in", y = "Total sequences (log)")
```

## ASV CLR value by supergroup

Get CLR transformed data for all ASVs

```{r}
head(insitu_asv_wClass)
```


```{r}
alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")
ciliate <- c("Ciliophora")

avg_insitu_asv <- insitu_asv_wClass %>% 
  # isolate euks only, remove opisthokonta
  filter(Domain == "Eukaryota") %>% 
  filter(Supergroup != "Opisthokonta") %>%
  # Modify supergroup level information
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  mutate(PHYLUM = case_when(
    Phylum == "Unknown" ~ paste(SUPERGROUP, "Other"),
    grepl("_X", Phylum) ~ paste(SUPERGROUP, "Other"),
    is.na(Phylum) ~ paste(SUPERGROUP, "Other"),
    TRUE ~ Phylum
  )) %>% 
  mutate(CLASS = case_when(
    Class == "Unknown" ~ PHYLUM,
    grepl("_X", Class) ~ PHYLUM,
    is.na(Class) ~ Phylum,
    grepl("MAST-", Class) ~ "MAST",
    TRUE ~ Class
  )) %>% 
  # filter(value > 200) %>% 
  # Average across replicates
  group_by(FeatureID, SITE, VENT, SAMPLETYPE, YEAR, Domain, SUPERGROUP, PHYLUM, CLASS, Order, Family, Genus, Species) %>% 
  summarise(AVG = mean(value)) %>% 
  ungroup() %>%  
  # isolate sample name and vent
  unite(SAMPLENAME_2, SITE, VENT, SAMPLETYPE, YEAR, sep = "_") %>% 
  pivot_wider(names_from = SAMPLENAME_2, values_from = AVG, values_fill = 0)
  ###
# names(avg_insitu_asv)
```

Perform CLR transformation
```{r}
# Isolate key for ASV ID and taxonomic information
tax_key <- select(avg_insitu_asv, FeatureID, Domain:Species)

# CLR, and re-format data frame
clr_trans_insitu_asv <- data.frame(compositions::clr(avg_insitu_asv %>% 
                               select(-(Domain:Species)) %>% 
                               column_to_rownames(var = "FeatureID"))) %>% 
  rownames_to_column(var = "FeatureID") %>% 
  pivot_longer(cols = starts_with(all), values_to = "CLR", names_to = "SAMPLENAME_2") %>%
  left_join(tax_key) %>% 
  separate(SAMPLENAME_2, c("SITE", "VENT", "SAMPLETYPE", "YEAR"), sep = "_") %>% 
  mutate(VENT = str_replace_all(VENT, "\\.", " ")) %>% 
      mutate(REGION = case_when(
        SITE == "GordaRidge" ~ "Gorda Ridge",
        SITE == "VonDamm" ~ "Von Damm",
        SITE == "Piccard" ~ "Piccard",
        SITE == "Axial" ~ "Axial")) %>% 
      mutate(VENTNAME = case_when(
        REGION == "Axial" ~ paste(VENT, YEAR, sep = " "),
        TRUE ~ VENT)) %>%
  unite(SAMPLE, REGION, VENTNAME, SAMPLETYPE, sep = " ", remove = FALSE) %>% 
  filter(CLR != 0) %>% 
  unite(sampletype_tax, SAMPLETYPE, SUPERGROUP, sep = " ", remove = FALSE) %>% 
  add_column(PRESENCE = 1)
# check output
# head(clr_trans_insitu_asv)
# hist(clr_trans_insitu_asv$CLR)
```

```{r, fig.height=6, fig.width=6}
sites_order <- c("Axial", "GordaRidge", "Piccard", "VonDamm")
clr_trans_insitu_asv$SITE_ORDER <- factor(clr_trans_insitu_asv$SITE, levels = rev(sites_order))
##
##

clr_trans_insitu_asv %>% 
  mutate(SUPERGROUP_ORDER = factor(SUPERGROUP, levels = all_taxa_order)) %>% 
  group_by(SAMPLETYPE, SUPERGROUP_ORDER, SITE_ORDER) %>% 
    summarize(COUNT_SUM = sum(PRESENCE))  %>% 
ggplot(aes(x = SAMPLETYPE, y = SITE_ORDER, fill = SUPERGROUP_ORDER)) +
  geom_point(shape = 21, aes(size = COUNT_SUM), stat = "identity") +
  facet_wrap(SUPERGROUP_ORDER ~ .) +
  scale_size_continuous(range = c(1,10)) +
  theme(legend.position = "top", 
        panel.grid.major = element_line(),
        panel.grid.minor = element_line(),
        panel.border = element_blank(), 
        panel.background = element_blank(),
        axis.line = element_line(),
        axis.text = element_text(color = "black", size = 8),
        strip.background = element_blank(), 
        strip.text = element_blank(),
        legend.title = element_blank(),
        axis.ticks = element_blank(),
        strip.placement = "outside") + 
  labs(x = "", y = "") +
  scale_fill_manual(values = all_taxa_color)
```


# Taxonomic diversity

## Tree map
Filter data to reduce noise and show sample type to vent ecosystem variability.
```{r, fig.height=7, fig.width=8}
alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")
bkgd <- c("Deep seawater", "BSW", "Shallow seawater", "Near vent BW")
plume <- c("Candelabra Plume", "Mt Edwards Plume", "Plume", "Anemone Plume")

to_supergroup <- insitu_asv_wClass %>% 
  filter(SITE %in% all) %>% 
  filter(Domain == "Eukaryota") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  mutate(SAMPLETYPEORDER = case_when(
    VENT %in% bkgd ~ "Background",
    VENT %in% plume ~ "Plume",
    TRUE ~ "Vent"
  )) %>% 
  group_by(FeatureID, Taxon, SUPERGROUP,
           VENT, SITE, SAMPLETYPE, YEAR, DATASET, SAMPLETYPEORDER) %>% 
    summarise(SEQ_AVG_REP = mean(value)) %>% 
  ungroup() %>% 
  group_by(SITE, SUPERGROUP, SAMPLETYPEORDER) %>% 
    summarise(ASV_COUNT = n(),
      SEQ_SUM = sum(SEQ_AVG_REP))
  
# Order sample type
to_supergroup$SAMPLETYPEORDER <- factor(to_supergroup$SAMPLETYPEORDER, levels = c("Background", "Plume", "Vent"))

# Remove ASVs with fewer than 200 sequences
# svg("../../../Manuscripts_presentations_reviews/deepsea-microeuk-ME/vectorgraphic-figures/fig2a-supergroup-treemap.svg", h = 7, w = 8)
to_supergroup %>% 
  mutate(SUPERGROUP_ORDER = factor(SUPERGROUP, levels = all_taxa_order)) %>% 
  filter(SEQ_SUM > 200) %>% 
  ggplot(aes(area = SEQ_SUM, fill = SUPERGROUP_ORDER, subgroup = SUPERGROUP)) +
    geom_treemap(color = "white") +
    geom_treemap_subgroup_border(colour = "white", size = 2) +
    facet_grid(SITE ~ SAMPLETYPEORDER) +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold", size = 12),
        strip.background = element_blank(), strip.text = element_text(color = "black", face = "bold", vjust = 0.5, hjust = 0, size = 12),
        legend.position = "right",
        legend.title = element_blank(),
        legend.text = element_text(color = "black", size = 12),
        panel.border = element_blank(),
        axis.title = element_text(face = "bold", size = 12)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_manual(values = all_taxa_color) +
  labs(x = "", y = "Sequence proportion by Supergroup")
# dev.off()
```


### Stats on protist supergroups, ASV & sequences

```{r}
# head(to_supergroup)
totalsum_supergroup <- sum(to_supergroup$SEQ_SUM)

table_perc_supergroup <- to_supergroup %>% 
  group_by(SITE) %>% 
  mutate(SUM_SITE = sum(SEQ_SUM)) %>%
  ungroup() %>% 
  group_by(SITE, SAMPLETYPEORDER) %>% 
  mutate(SUM_SITE_TYPE = sum(SEQ_SUM)) %>% 
  ungroup() %>% 
mutate(Perc_seq_total = (100*(SEQ_SUM/totalsum_supergroup)),
       Perc_seq_site = (100*(SEQ_SUM/SUM_SITE)),
       Perc_seq_site_type = (100*(SEQ_SUM/SUM_SITE_TYPE))) %>% 
  select(-starts_with("SUM_")) %>% 
  gt(
    groupname_col = c("SITE", "SAMPLETYPEORDER")
    # rowname_col = "SAMPLENAME"
  ) %>% 
  fmt_number(columns = starts_with("Perc_"), decimals = 2) %>% 
  fmt_number(columns = SEQ_SUM, decimals = 0) %>% 
  fmt_number(columns = ASV_COUNT, decimals = 0)

# gtsave(table_perc_supergroup, filename = "output-tables/perc_seq_asv_bysupergroup.html")

```


# Distance-decay

## Calculate distances across all samples
```{r}
library(geosphere)

# Attach better coordinates (fixed in 'metadata-compilation-table.R')
coords_input <- read.delim("tableS1-geochem-params.txt") %>% 
  select(VENT, SITE, SAMPLETYPE, LONG, LAT) %>% 
  unite(SAMPLENAME, VENT, SITE, SAMPLETYPE, sep = "_") %>% 
  distinct(SAMPLENAME, .keep_all = TRUE) %>%
  column_to_rownames(var = "SAMPLENAME")

vents_all <- row.names(coords_input)
distances_m <- as.data.frame(distm(coords_input, fun = distHaversine))
colnames(distances_m) <- vents_all
rownames(distances_m) <- vents_all

# Distinction with "_all" here is that some samples are not included in below analysis. 
```


Calculate distances and create long format dataframe
```{r}
dist_m <- distances_m %>% 
  rownames_to_column(var = "start") %>% 
  pivot_longer(cols = vents_all, names_to = "end") %>%
  filter(!(start == end))
```


## Format input ASV data
```{r}
# Widen an numeric the ASVs for the distance calculations
# vents # use this list of samples

df_wide_num <- insitu_asv_wClass %>% 
  group_by(FeatureID, SAMPLENAME, VENT, SITE, SAMPLETYPE) %>% 
  summarise(AVG = mean(value)) %>% 
    ungroup() %>%
  select(-SAMPLENAME) %>% 
  unite(SAMPLENAME, VENT, SITE, SAMPLETYPE, sep = "_") %>% 
  group_by(FeatureID, SAMPLENAME) %>% 
  summarise(SUM = sum(AVG)) %>% 
  pivot_wider(names_from = FeatureID, values_from = SUM, values_fill = 0) %>% 
  column_to_rownames(var = "SAMPLENAME")

vents <- row.names(df_wide_num)
```


## Calculate distance matrices

Inclusion of three metrics below is because I want to see if the answer varies across the 
```{r}
dist_bray <- as.data.frame(as.matrix(vegdist(df_wide_num, method = "bray"))) %>% 
  rownames_to_column(var = "start") %>% 
  pivot_longer(cols = vents, names_to = "end") %>% 
  select(start, end, Bray_Curtis_metric = value)

dist_jacc <- as.data.frame(as.matrix(vegdist(df_wide_num, method = "jaccard"))) %>% 
  rownames_to_column(var = "start") %>% 
  pivot_longer(cols = vents, names_to = "end") %>% 
  select(start, end, Jaccard_metric = value)

dist_euc_clr <- as.data.frame(as.matrix(vegdist(compositions::clr(df_wide_num), method = "euclidean"))) %>% 
  rownames_to_column(var = "start") %>% 
  pivot_longer(cols = vents, names_to = "end") %>% 
  select(start, end, Euclidean_CLR_metric = value)
```


## Compile distances of pairwise comparisons
```{r}
dist_compiled_output <- dist_m %>% 
  select(start, end, meters = value) %>% 
  left_join(dist_jacc) %>% 
  left_join(dist_bray) %>% 
  left_join(dist_euc_clr) %>% 
  pivot_longer(cols = ends_with("_metric"), names_to = "comm_dist") %>% 
  filter(!(start == end)) %>%
  mutate(km = (meters/1000),
         dist0 = gsub("_", " ", comm_dist),
         dist = gsub(" metric", "", dist0))
  
```

## Plot distance-decay

Compare several distance metrics with distance.
```{r, fig.height=7, fig.width=4}
# ggplot(dist_compiled_output, aes(x = km, y = value)) +
#   geom_jitter() +
#   scale_x_log10(labels = scales::comma) +
#   facet_grid(dist ~ ., scales = "free") +
#   theme_linedraw() +
#   geom_smooth(method="lm") +
#   stat_regline_equation(aes(label = ..rr.label..), color = "black", label.y.npc = 0.09, label.x.npc = 0.7) +
#   theme(strip.background = element_blank(),
#         strip.text = element_text(color = "black")) +
#   labs(y = "Distance metric")
```

```{r, fig.height=4, fig.width=5}
# svg("../../../Manuscripts_presentations_reviews/deepsea-microeuk-ME/vectorgraphic-figures/fig4a-distdecay.svg", h = 4, w = 5)
dist_compiled_output %>% 
  filter(dist == "Jaccard") %>% 
  ggplot(aes(x = km, y = value)) +
  geom_jitter() +
  scale_x_log10(labels = scales::comma) +
  facet_grid(dist ~ ., scales = "free") +
  # geom_smooth(method="lm") +
  theme_linedraw() +
  # stat_regline_equation(aes(label = ..rr.label..), face = "bold", color = "black", label.y.npc = 0.09, label.x.npc = 0.7) +
  scale_y_continuous(limits = c(0.7,1)) +
  theme(strip.background = element_blank(),
        strip.text = element_text(color = "black", face = "bold", size = 12),
        axis.text = element_text(color = "black", face = "bold", size = 12),
        axis.title = element_text(color = "black", face = "bold", size = 12)) +
  labs(y = "Jaccard Dissimilarity")
# dev.off()
# labs(y = "Distance metric", title = "Distance-Decay: all ASVs")
```

## Plot variance of Jaccard index by distance in km
```{r, fig.height=4, fig.width=5}
# head(dist_compiled_output)
# # ?var()
# svg("../../../Manuscripts_presentations_reviews/deepsea-microeuk-ME/vectorgraphic-figures/figS4-var-dist.svg", h = 4, w = 5)
var_km <- dist_compiled_output %>% 
  filter(dist == "Jaccard") %>% 
  # separate(start, into = c("NAME_start", "SITE_start"), sep = "_", remove = FALSE, extra = "merge") %>%
  # separate(end, into = c("NAME_end", "SITE_end"), sep = "_", remove = FALSE, extra = "merge") %>% 
  separate(start, into = c("NAME_start", "SITE_start", "TYPE_start"), sep = "_", remove = FALSE, extra = "merge") %>%
  separate(end, into = c("NAME_end", "SITE_end", "TYPE_end"), sep = "_", remove = FALSE, extra = "merge") %>%
  drop_na() %>% 
  # group_by(start, end) %>% 
  group_by(SITE_start, SITE_end) %>%
  mutate(VARIANCE = var(value)) %>% 
ggplot(aes(x = km, y = VARIANCE)) +
  scale_x_log10(labels = scales::comma) +
  geom_jitter(shape = 21, size = 2, color = "black", fill = "black") +
  geom_smooth(method="lm") +
  theme_linedraw() +
  stat_regline_equation(aes(label = ..rr.label..), color = "black", label.y.npc = 0.0007, label.x.npc = 0.08) +
  # scale_y_continuous(limits = c(0.7,1)) +
  theme(strip.background = element_blank(),
        strip.text = element_text(color = "black"),
        axis.text = element_text(color = "black"),
        axis.title = element_text(color = "black")) +
  labs(y = "Variance of Jaccard Dissimilarity", x = "")
# dev.off()
```


## Repeat distance-decay with vent-only sites

Sample to vent sites only
```{r}
# head(insitu_asv_wClass)
df_wide_vent <- insitu_asv_wClass %>% 
  filter(SAMPLETYPE == "Vent") %>% 
  group_by(FeatureID, SAMPLENAME, VENT, SITE, SAMPLETYPE) %>% 
  summarise(AVG = mean(value)) %>% 
    ungroup() %>%
  select(-SAMPLENAME) %>% 
  unite(SAMPLENAME, VENT, SITE, SAMPLETYPE, sep = "_") %>% 
  group_by(FeatureID, SAMPLENAME) %>% 
  summarise(SUM = sum(AVG)) %>% 
  pivot_wider(names_from = FeatureID, values_from = SUM, values_fill = 0) %>% 
  column_to_rownames(var = "SAMPLENAME")

vents <- rownames(df_wide_vent)
```


```{r}
dist_bray <- as.data.frame(as.matrix(vegdist(df_wide_vent, method = "bray"))) %>% 
  rownames_to_column(var = "start") %>% 
  pivot_longer(cols = vents, names_to = "end") %>% 
  select(start, end, Bray_Curtis_metric = value)

dist_jacc <- as.data.frame(as.matrix(vegdist(df_wide_vent, method = "jaccard"))) %>% 
  rownames_to_column(var = "start") %>% 
  pivot_longer(cols = vents, names_to = "end") %>% 
  select(start, end, Jaccard_metric = value)

dist_euc_clr <- as.data.frame(as.matrix(vegdist(compositions::clr(df_wide_vent), method = "euclidean"))) %>% 
  rownames_to_column(var = "start") %>% 
  pivot_longer(cols = vents, names_to = "end") %>% 
  select(start, end, Euclidean_CLR_metric = value)

dist_compiled_output_vents <- dist_m %>% 
  select(start, end, meters = value) %>% 
  left_join(dist_jacc) %>% 
  left_join(dist_bray) %>% 
  left_join(dist_euc_clr) %>% 
  pivot_longer(cols = ends_with("_metric"), names_to = "comm_dist") %>% 
  filter(!(start == end)) %>%
  mutate(km = (meters/1000),
         dist0 = gsub("_", " ", comm_dist),
         dist = gsub(" metric", "", dist0))
```


```{r, fig.height=7, fig.width=4}
dist_compiled_output_vents %>% 
  # filter(grepl("Vent", start) & grepl("Vent", end)) %>%
  ggplot(aes(x = km, y = value)) +
  geom_jitter() +
  scale_x_log10(labels = scales::comma) +
  facet_grid(dist ~ ., scales = "free") +
  geom_smooth(method="lm") +
  theme_linedraw() +
  stat_regline_equation(aes(label = ..rr.label..), color = "black", label.y.npc = 0.09, label.x.npc = 0.7) +
  theme(strip.background = element_blank(),
        strip.text = element_text(color = "black")) +
  labs(y = "Distance metric")
```

```{r, fig.height=4, fig.width=5}
vent_dist_decay <- dist_compiled_output_vents %>% 
  filter(dist == "Jaccard") %>% 
  ggplot(aes(x = km, y = value)) +
  geom_jitter() +
  scale_x_log10(labels = scales::comma) +
  facet_grid(dist ~ ., scales = "free") +
  geom_smooth(method="lm") +
  theme_linedraw() +
  stat_regline_equation(aes(label = ..rr.label..), color = "black", label.y.npc = 0.09, label.x.npc = 0.7) +
  theme(strip.background = element_blank(),
        strip.text = element_text(color = "black")) +
  labs(y = "Jaccard Dissimilarity", title = "",  x = "")
```


## Repeat distance-decay with subpopulations
Sample to resident only and cosmopolitan only
```{r}
cosmo <- c("Vent, plume, & background", "Vent & background", "Vent & plume", "Plume & background")
res <- c("Vent only")

df_wide_res <- insitu_asv_wClass %>% 
  filter(CLASS %in% res) %>% 
  group_by(FeatureID, SAMPLENAME, VENT, SITE, SAMPLETYPE) %>% 
  summarise(AVG = mean(value)) %>% 
    ungroup() %>%
  select(-SAMPLENAME) %>% 
  unite(SAMPLENAME, VENT, SITE, SAMPLETYPE, sep = "_") %>% 
  group_by(FeatureID, SAMPLENAME) %>% 
  summarise(SUM = sum(AVG)) %>% 
  pivot_wider(names_from = FeatureID, values_from = SUM, values_fill = 0) %>% 
  column_to_rownames(var = "SAMPLENAME")
#
vents_res <- row.names(df_wide_res)
#
```


```{r}
dist_bray <- as.data.frame(as.matrix(vegdist(df_wide_res, method = "bray"))) %>% 
  rownames_to_column(var = "start") %>% 
  pivot_longer(cols = vents_res, names_to = "end") %>% 
  select(start, end, Bray_Curtis_metric = value)

dist_jacc <- as.data.frame(as.matrix(vegdist(df_wide_res, method = "jaccard"))) %>% 
  rownames_to_column(var = "start") %>% 
  pivot_longer(cols = vents_res, names_to = "end") %>% 
  select(start, end, Jaccard_metric = value)

dist_euc_clr <- as.data.frame(as.matrix(vegdist(compositions::clr(df_wide_res), method = "euclidean"))) %>% 
  rownames_to_column(var = "start") %>% 
  pivot_longer(cols = vents_res, names_to = "end") %>% 
  select(start, end, Euclidean_CLR_metric = value)

dist_compiled_output_res <- dist_m %>% 
  select(start, end, meters = value) %>% 
  left_join(dist_jacc) %>% 
  left_join(dist_bray) %>% 
  left_join(dist_euc_clr) %>% 
  pivot_longer(cols = ends_with("_metric"), names_to = "comm_dist") %>% 
  filter(!(start == end)) %>%
  mutate(km = (meters/1000),
         dist0 = gsub("_", " ", comm_dist),
         dist = gsub(" metric", "", dist0))
```

```{r, fig.height=4, fig.width=5}
d_d_resident <- dist_compiled_output_res %>% 
  filter(dist == "Jaccard") %>% 
  ggplot(aes(x = km, y = value)) +
  geom_jitter() +
  scale_x_log10(labels = scales::comma) +
  facet_grid(dist ~ ., scales = "free") +
  geom_smooth(method="lm") +
  theme_linedraw() +
  stat_regline_equation(aes(label = ..rr.label..), color = "black", label.y.npc = 0.09, label.x.npc = 0.7) +
  theme(strip.background = element_blank(),
        strip.text = element_text(color = "black")) +
  labs(y = "Distance metric", title = "Distance-Decay: Resident ASVs")
d_d_resident
```

Target cosmpolitan
```{r}
cosmo <- c("Vent, plume, & background", "Vent & background", "Vent & plume", "Plume & background")
res <- c("Vent only")

df_wide_cosmo <- insitu_asv_wClass %>% 
  filter(CLASS %in% cosmo) %>% 
  group_by(FeatureID, SAMPLENAME, VENT, SITE, SAMPLETYPE) %>% 
  summarise(AVG = mean(value)) %>% 
    ungroup() %>%
  select(-SAMPLENAME) %>% 
  unite(SAMPLENAME, VENT, SITE, SAMPLETYPE, sep = "_") %>% 
  group_by(FeatureID, SAMPLENAME) %>% 
  summarise(SUM = sum(AVG)) %>% 
  pivot_wider(names_from = FeatureID, values_from = SUM, values_fill = 0) %>% 
  column_to_rownames(var = "SAMPLENAME")
#
vents_cosmo <- row.names(df_wide_cosmo)
#
```


```{r}
dist_bray <- as.data.frame(as.matrix(vegdist(df_wide_cosmo, method = "bray"))) %>% 
  rownames_to_column(var = "start") %>% 
  pivot_longer(cols = vents_cosmo, names_to = "end") %>% 
  select(start, end, Bray_Curtis_metric = value)

dist_jacc <- as.data.frame(as.matrix(vegdist(df_wide_cosmo, method = "jaccard"))) %>% 
  rownames_to_column(var = "start") %>% 
  pivot_longer(cols = vents_cosmo, names_to = "end") %>% 
  select(start, end, Jaccard_metric = value)

dist_euc_clr <- as.data.frame(as.matrix(vegdist(compositions::clr(df_wide_cosmo), method = "euclidean"))) %>% 
  rownames_to_column(var = "start") %>% 
  pivot_longer(cols = vents_cosmo, names_to = "end") %>% 
  select(start, end, Euclidean_CLR_metric = value)

dist_compiled_output_cosmo<- dist_m %>% 
  select(start, end, meters = value) %>% 
  left_join(dist_jacc) %>% 
  left_join(dist_bray) %>% 
  left_join(dist_euc_clr) %>% 
  pivot_longer(cols = ends_with("_metric"), names_to = "comm_dist") %>% 
  filter(!(start == end)) %>%
  mutate(km = (meters/1000),
         dist0 = gsub("_", " ", comm_dist),
         dist = gsub(" metric", "", dist0))
```

```{r, fig.height=4, fig.width=5}
cosmo_dist_decay <- dist_compiled_output_cosmo %>% 
  filter(dist == "Jaccard") %>% 
  ggplot(aes(x = km, y = value)) +
  geom_jitter() +
  scale_x_log10(labels = scales::comma) +
  facet_grid(dist ~ ., scales = "free") +
  geom_smooth(method="lm") +
  theme_linedraw() +
  stat_regline_equation(aes(label = ..rr.label..), color = "black", label.y.npc = 0.09, label.x.npc = 0.7) +
  theme(strip.background = element_blank(),
        strip.text = element_text(color = "black")) +
  labs(y = "Jaccard Dissimilarity", title = "", x = "")
```

Figure S4
```{r, fig.height=10, fig.width=9}
# svg("../../../Manuscripts_presentations_reviews/deepsea-microeuk-ME/vectorgraphic-figures/supp_dist_decay_figS4a-c.svg", h = 10, w = 5)
# var_km + vent_dist_decay + cosmo_dist_decay +patchwork::plot_layout(nrow = 3) + plot_annotation(tag_levels = 'a')

# Execute all commands  in clustered-ASV-analysis markdown file to get dist_decay for OTU clustering
# svg("../../../Manuscripts_presentations_reviews/deepsea-microeuk-ME/vectorgraphic-figures/supp_dist_decay_figS4_all.svg", h = 10, w = 9)
# ((var_km / vent_dist_decay / cosmo_dist_decay + plot_layout(guides = 'keep')) | dist_decay) + plot_layout(guides = 'collect') + plot_annotation(tag_levels = 'a')
# dev.off()
# dev.off()
```



Isolate taxa key, and manually curated taxonomic groupings.
```{r}
alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")
ciliate <- c("Ciliophora")

tax_key_curated <- insitu_asv_wClass %>% 
  select(FeatureID, Taxon, Domain, Supergroup, Phylum, Class, Order, Family, Genus, Species) %>% 
  distinct() %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  mutate(PHYLUM = case_when(
    Phylum == "Unknown" ~ paste(SUPERGROUP, "Other"),
    grepl("_X", Phylum) ~ paste(SUPERGROUP, "Other"),
    is.na(Phylum) ~ paste(SUPERGROUP, "Other"),
    TRUE ~ Phylum
  )) %>% 
  mutate(CLASS = case_when(
    Class == "Unknown" ~ PHYLUM,
    grepl("_X", Class) ~ PHYLUM,
    is.na(Class) ~ Phylum,
    grepl("MAST-", Class) ~ "MAST",
    TRUE ~ Class
  )) %>% 
  mutate(CLASS2 = case_when(
    CLASS == Order ~ Order,
    is.na(Order) ~ paste(CLASS, "Other"),
    TRUE ~ Order)) %>%
  filter(!is.na(Order)) %>%
  select(FeatureID, SUPERGROUP, PHYLUM, CLASS, CLASS2) %>% 
  mutate(SUPERGROUP_ORDER = factor(SUPERGROUP, levels = all_taxa_order))
# head(tax_key_curated)
# unique(tax_key_curated$CLASS_2)
```


# Plot total sequences by # of vent sites

```{r}
tmp <- insitu_asv_wClass %>%
    filter(Domain == "Eukaryota") %>% 
    # filter(CLASS %in% slice) %>%
    # Average across replicates
      group_by(FeatureID, SAMPLENAME, VENT, distribution = CLASS) %>% 
      summarise(AVG = mean(value)) %>% 
      ungroup() %>% 
    left_join(tax_key_curated) %>% 
    separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp")) %>% 
    mutate(VENTNAME = case_when(
      SITE == "Axial" ~ paste(SITE, VENT, YEAR, sep = " "),
      SITE == "GordaRidge" ~ paste(SITE, VENT, sep = " "),
      SITE %in% mcr ~ paste(SITE, VENT, sep = " ")
      )) %>% select(-Sample_tmp) %>% 
  # filter(SAMPLETYPE == "Vent") %>% 
  group_by(VENTNAME, distribution, SUPERGROUP, PHYLUM, CLASS2) %>% 
    summarise(TOTAL_SEQ = sum(AVG),
              COUNT = n()) # %>% #IF you use this please modify location
  # filter(!is.na(SUPERGROUP))

head(tmp)
```
```{r, fig.width=18, fig.height=10}
tmp %>% 
  filter(distribution == "Vent only") %>% 
  ggplot(aes(x = VENTNAME, y = CLASS2)) +
  geom_point(aes(fill = log(TOTAL_SEQ), size = COUNT), shape = 21) +
  scale_size_continuous(range = c(1,18)) +
  # scale_shape_manual(values = c(21, 23, 22, 25)) +
  facet_grid(.~SUPERGROUP, space = "free", scales = "free") +
  theme(axis.text.x = element_text(angle = 90)) +
  coord_flip()
```


# Pairwise comparison of community diversity

## Compare resident vs.cosmopolitan population
```{r}
cosmo <- c("Vent, plume, & background", "Vent & background", "Vent & plume", "Plume & background")
res <- c("Vent only")
all_classes <- as.character(unique(insitu_asv_wClass$CLASS))

axial <- c("Axial")
mcr <- c("VonDamm", "Piccard")
gr <- c("GordaRidge")
all <- c("Axial", "VonDamm", "Piccard", "GordaRidge")
```


## Jaccard comparison of resident vs. cosmopolitan
```{r, fig.height=8, fig.width=8}
isolate_calc_jaccard <- function(df, slice){
  df_out_wide <- df %>%
    filter(Domain == "Eukaryota") %>% 
    filter(CLASS %in% slice) %>%
    # Average across replicates
      group_by(FeatureID, SAMPLENAME, VENT) %>% 
      summarise(AVG = mean(value)) %>% 
      ungroup() %>% 
    separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp")) %>% 
    mutate(VENTNAME = case_when(
      SITE == "Axial" ~ paste(SITE, VENT, YEAR, sep = " "),
      SITE == "GordaRidge" ~ paste(SITE, VENT, sep = " "),
      SITE %in% mcr ~ paste(SITE, VENT, sep = " ")
      )) %>% select(-Sample_tmp) %>% 
    unite(SAMPLEID, SITE, VENT, YEAR, SAMPLETYPE, sep = "-") %>%
    unite(sample_full, SAMPLEID, VENTNAME, sep = "--") %>% 
     group_by(FeatureID, sample_full) %>% 
      summarise(SUM = sum(AVG)) %>% 
    pivot_wider(names_from = FeatureID, values_from = SUM, values_fill = 0) %>% 
    column_to_rownames(var = "sample_full")
  }
```
Apply function to resident population, cosmopolitan population, and whole population
```{r}
resident <- isolate_calc_jaccard(insitu_asv_wClass, res)
cosmopolitan <- isolate_calc_jaccard(insitu_asv_wClass, cosmo)
all_jacc <- isolate_calc_jaccard(insitu_asv_wClass, all_classes)
# head(resident)
```


Plot resident
```{r}
dist_jacc_clr_res <- as.data.frame(as.matrix(vegdist(compositions::clr(resident), method = "euclidean"))) %>%
    rownames_to_column(var = "start") %>% 
    pivot_longer(cols = row.names(resident), names_to = "end") %>% 
    select(start, end, Jaccard_CLR_metric = value) %>% 
    # mutate(VARS = purrr::map2_chr(start, end, ~toString(sort(c(.x, .y))))) %>%
    # distinct(VARS, .keep_all = TRUE) %>%
    # separate(VARS, c("X", "Y"), sep = ", ") %>% 
    # select(-start, -end) %>% 
    # add_column(SUBSET = "RESIDENT") %>% 
    separate(start, c("Y_FULL", "Y"), sep = "--", remove = FALSE) %>% 
    separate(Y_FULL, c("Y_SITE", "Y_VENT", "Y_YEAR"), sep = "-", remove = FALSE) %>% 
    separate(end, c("X_FULL", "X"), sep = "--", remove = FALSE) %>% 
    separate(X_FULL, c("X_SITE", "X_VENT", "X_YEAR"), sep = "-", remove = FALSE) %>% 
  ggplot(aes(x = X, y = Y, fill = Jaccard_CLR_metric)) +
  geom_tile(color = "black") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.title = element_blank()) +
  # scale_fill_gradientn(colors = c("#badbdb", "#dedad2", "#e4bcad", "#df979e", "#d7658b", "#c80064")) +
  # scale_fill_gradientn(colors = c("#fff7f3", "#fde0dd", "#fcc5c0", "#fa9fb5", "#f768a1", "#dd3497", "#ae017e", "#7a0177", "#49006a"))
  scale_fill_gradientn(colors = c("#ffffff","#f0f0f0","#d9d9d9","#bdbdbd","#969696","#737373","#525252","#252525","#000000")) +
  coord_fixed(ratio = 1)
dist_jacc_clr_res
```

Plot cosmopolitan
```{r, fig.height=8}
dist_jacc_clr_cos <- as.data.frame(as.matrix(vegdist(compositions::clr(cosmopolitan), method = "euclidean"))) %>%
    rownames_to_column(var = "start") %>% 
    pivot_longer(cols = row.names(cosmopolitan), names_to = "end") %>% 
    select(start, end, Jaccard_CLR_metric = value) %>% 
    separate(start, c("Y_FULL", "Y"), sep = "--", remove = FALSE) %>% 
    separate(Y_FULL, c("Y_SITE", "Y_VENT", "Y_YEAR"), sep = "-", remove = FALSE) %>% 
    separate(end, c("X_FULL", "X"), sep = "--", remove = FALSE) %>% 
    separate(X_FULL, c("X_SITE", "X_VENT", "X_YEAR"), sep = "-", remove = FALSE) %>% 
  ggplot(aes(x = X, y = Y, fill = Jaccard_CLR_metric)) +
  geom_tile(color = "black") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.title = element_blank()) +
  # scale_fill_gradientn(colors = c("#badbdb", "#dedad2", "#e4bcad", "#df979e", "#d7658b", "#c80064")) +
  # scale_fill_gradientn(colors = c("#fff7f3", "#fde0dd", "#fcc5c0", "#fa9fb5", "#f768a1", "#dd3497", "#ae017e", "#7a0177", "#49006a"))
  scale_fill_gradientn(colors = c("#ffffff","#f0f0f0","#d9d9d9","#bdbdbd","#969696","#737373","#525252","#252525","#000000")) +
  coord_fixed(ratio = 1)
dist_jacc_clr_cos
```

Plot all ASVs
```{r}
dist_jacc_clr_all <- as.data.frame(as.matrix(vegdist(compositions::clr(all_jacc), method = "euclidean"))) %>%
    rownames_to_column(var = "start") %>% 
    pivot_longer(cols = row.names(all_jacc), names_to = "end") %>% 
    select(start, end, Jaccard_CLR_metric = value) %>% 
    separate(start, c("Y_FULL", "Y"), sep = "--", remove = FALSE) %>% 
    separate(Y_FULL, c("Y_SITE", "Y_VENT", "Y_YEAR"), sep = "-", remove = FALSE) %>% 
    separate(end, c("X_FULL", "X"), sep = "--", remove = FALSE) %>% 
    separate(X_FULL, c("X_SITE", "X_VENT", "X_YEAR"), sep = "-", remove = FALSE) %>% 
  ggplot(aes(x = X, y = Y, fill = Jaccard_CLR_metric)) +
  geom_tile(color = "black") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.title = element_blank()) +
  # scale_fill_gradientn(colors = c("#badbdb", "#dedad2", "#e4bcad", "#df979e", "#d7658b", "#c80064")) +
  # scale_fill_gradientn(colors = c("#fff7f3", "#fde0dd", "#fcc5c0", "#fa9fb5", "#f768a1", "#dd3497", "#ae017e", "#7a0177", "#49006a"))
  scale_fill_gradientn(colors = c("#ffffff","#f0f0f0","#d9d9d9","#bdbdbd","#969696","#737373","#525252","#252525","#000000")) +
  coord_fixed(ratio = 1)
dist_jacc_clr_all
```



# Plot population sets as CLR

```{r}
clr_to_plot_maintax <- function(df, slice, tax_slice){
  df_out_wide <- df %>%
    filter(Domain == "Eukaryota") %>%
    filter(CLASS %in% slice) %>%
    # Average across replicates
      group_by(FeatureID, SAMPLENAME, VENT) %>% 
      summarise(AVG = mean(value)) %>% 
      ungroup() %>% 
    separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp")) %>% 
    mutate(VENTNAME = case_when(
      SITE == "Axial" ~ paste(SITE, VENT, YEAR, sep = " "),
      SITE == "GordaRidge" ~ paste(SITE, VENT, sep = " "),
      SITE %in% mcr ~ paste(SITE, VENT, sep = " ")
      )) %>% select(-Sample_tmp) %>% 
    left_join(tax_key_curated) %>%
    # Modify Class 2 level with excess text.
    mutate(CLASS2 = str_replace_all(CLASS2, "_X", "")) %>%
    mutate(CLASS2 = str_replace_all(CLASS2, "_XX", "")) %>%
    mutate(CLASS2 = str_replace_all(CLASS2, "X$", "")) %>% 
    mutate(CLASS2 = str_replace_all(CLASS2, "_[:alnum:]$", "")) %>% 
    group_by(VENTNAME, SUPERGROUP, PHYLUM, CLASS, CLASS2) %>% 
        summarise(SUM_VALUE = sum(AVG)) %>% 
    filter(SUPERGROUP %in% tax_slice) %>% 
    unite(TAX, SUPERGROUP, PHYLUM, CLASS, CLASS2, sep = "__") %>% 
    pivot_wider(names_from = TAX, values_from = SUM_VALUE, values_fill = 0) %>% 
    column_to_rownames(var = "VENTNAME")
  # Plot
  as.data.frame(compositions::clr(df_out_wide)) %>% 
  rownames_to_column(var = "VENTNAME") %>%
    pivot_longer(cols = names(df_out_wide), values_to = "CLR", names_to = "TAX") %>% 
    separate(TAX, into = c("SUPERGROUP", "PHYLUM", "CLASS", "CLASS2"), sep = "__") %>% 
    separate(VENTNAME, into = c("SITE"), sep = " ", remove = FALSE) %>% 
    ggplot(aes(x = CLASS2, y = VENTNAME, fill = CLR)) +
             geom_tile(aes(fill = CLR), color = "black") +
      theme(legend.position = "right", 
            panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(), 
            panel.border = element_blank(), 
            panel.background = element_blank(),
            axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, color = "black",size = 10, face = "bold"), 
            axis.text.y = element_text(color = "black", size = 10, face = "bold", angle = 45, hjust = 1, vjust = 1),
            strip.background = element_blank(), 
            strip.text.y = element_text(hjust = 0, vjust = 0, angle = 45, size = 10, color = "black", face = "bold"),
            strip.text.x = element_text(hjust = 1, vjust = 0, angle = 90, size = 10, color = "black", face = "bold"),
            legend.title = element_blank(),
            strip.placement = "outside",
            axis.ticks.y = element_blank()) + 
      labs(x = "", y = "") +
    scale_fill_steps2(
  low = "#2166ac",
  mid = "white",
  high = "#b2182b",
  midpoint = 0,
  space = "Lab",
  na.value = "#4d4d4d",
  guide = "coloursteps",
  aesthetics = "fill") +
    # coord_flip() +
    facet_grid(SITE ~ SUPERGROUP + PHYLUM, space = "free", scales = "free", switch = "both")
}
```

```{r, fig.height=11, fig.width=21}
# svg("../../../Manuscripts_presentations_reviews/deepsea-microeuk-ME/vectorgraphic-figures/fig5b-clr-tile.svg", h = 11, w = 21)
maintext_supergroup <- c("Alveolata-Ciliophora", "Amoebozoa", "Apusozoa", "Excavata", "Rhizaria-Cercozoa", "Rhizaria-Radiolaria")
clr_to_plot_maintax(insitu_asv_wClass, res, maintext_supergroup)
# dev.off()
```

```{r, fig.width=10, fig.height=6}
unique(tax_key_curated$SUPERGROUP)
mast <- c("Stramenopiles-Opalozoa", "Stramenopiles-Sagenista")
clr_to_plot_maintax(insitu_asv_wClass, res, mast)
```


```{r, fig.height=6, fig.width=17}
# maintext_supergroup <- c("Alveolata-Ciliophora", "Alveolata-Dinoflagellata", "Rhizaria-Cercozoa", "Rhizaria-Radiolaria")
# clr_to_plot_maintax(insitu_asv_wClass, res, maintext_supergroup)
```


```{r, fig.height=6, fig.width=17}
# maintext_supergroup <- c("Alveolata-Ciliophora", "Other Alveolata", "Amoebozoa", "Apusozoa", "Rhizaria-Cercozoa", "Rhizaria-Radiolaria")
# clr_to_plot_maintax(insitu_asv_wClass, all_classes, maintext_supergroup)
```

# Driving forces in microeukaryote community structure

## Diversity estimators (DivNet)
Run on HPC at WHOI.

```{r Load from local, eval=FALSE}
load("asv-tables-processed-18102021.RData", verbose = T)
```

DivNet package - diversity estimation hypothesis testing from Amy
Willis's group. This will also characterize the uncertainty of the
richness estimate. Richness estimation is flawed because of sample depth and processing methods.

```{r load relevant libraries for diversity estimations}
library(phyloseq); library(breakaway); library(DivNet)
library(tidyverse)
```

This code block run on HPC.

```{r create phyloseq objects for divnet}
# Select eukaryotes only and create wide format dataframe
insitu_wide <- asv_insitu_qc %>% 
  filter(Domain == "Eukaryota") %>% 
  filter(!grepl("_Plume001_", SAMPLE)) %>% #removing "near vent background", not relevant in other data sets
  select(FeatureID, Taxon, SAMPLE, value) %>%
  pivot_wider(names_from = SAMPLE, values_from = value, values_fill = 0)

# head(insitu_wide)
insitu_samples <- as.character(colnames(insitu_wide %>% select(-Taxon, -FeatureID)))
# insitu_samples
```

```{r Isolate matrix for asvs and taxa}
insitu_tax_matrix <- insitu_wide %>%
  select(FeatureID, Taxon) %>% 
  separate(Taxon, c("Domain", "Supergroup", 
                  "Phylum", "Class", "Order",
                  "Family", "Genus", "Species"), sep = ";") %>% 
  column_to_rownames(var = "FeatureID") %>% 
  as.matrix

insitu_asv_matrix <- insitu_wide %>% 
  select(-Taxon) %>% 
  column_to_rownames(var = "FeatureID") %>% 
  as.matrix

# Align row names for each matrix
rownames(insitu_tax_matrix) <- row.names(insitu_asv_matrix)
```

Reformat metadata, but change to numeric.
```{r}
metadata <- read.csv("data-input/samplelist-metadata.csv", na.strings = "NA")
# ?read.csv()
## Extract relevant metadata information
metadata_formatted <- metadata %>% 
  mutate_all(as.character) %>%
  filter(Sample_or_Control == "Sample") %>% 
  filter(!(SAMPLETYPE == "Incubation")) %>% 
  filter(!(SAMPLETYPE == "Microcolonizer")) %>% 
  select(SAMPLE, VENT, COORDINATES, SITE, SAMPLEID, DEPTH, SAMPLETYPE, YEAR, TEMP = starts_with("TEMP"), pH, PercSeawater = starts_with("Perc"), Mg = starts_with("Mg"), H2 = starts_with("H2."), H2S = starts_with("H2S"), CH4 = starts_with("CH4"), ProkConc, Sample_or_Control)

# Ensure these columns are converted to numeric
env_params <- c("DEPTH", "TEMP", "pH", "PercSeawater", "Mg", "H2", "H2S", "CH4", "ProkConc")

metadata_insitu <- metadata_formatted %>%
  na_if(., "bd") %>% na_if(., "nd") %>% na_if(., "-") %>% na_if(., "") %>% na_if(., "na") %>%
  filter(SAMPLE %in% insitu_samples) %>% # from reformatting df above
  # select(SAMPLE, VENT, SITE, SAMPLETYPE, YEAR) %>%
  unite(SAMPLELABEL, VENT, SITE, SAMPLETYPE, YEAR, sep = "_", remove = FALSE) %>% 
  unite(TYPE_SITE, SITE, SAMPLETYPE, sep = "_", remove = FALSE) %>% 
  mutate_at(env_params,as.numeric)

rownames(metadata_insitu) <- metadata_insitu$SAMPLE
# str(metadata_insitu)
# View(metadata_insitu)
# head(metadata_insitu)
# row.names(metadata_insitu)
```


Import taxa and ASV count matrices into phyloseq objects.

```{r Phyloseq import contined}
# Import asv and tax matrices
ASV = otu_table(insitu_asv_matrix, taxa_are_rows = TRUE)
TAX = tax_table(insitu_tax_matrix)

phylo_obj <- phyloseq(ASV, TAX)

# Import metadata as sample data in phyloseq
samplenames <- sample_data(metadata_insitu)

# join as phyloseq object
physeq_insitu = merge_phyloseq(phylo_obj, samplenames)

## Check
physeq_insitu
# head(insitu_tax_matrix)
# head(metadata_insitu)

save(physeq_insitu, metadata_insitu, file = "phyloseq-objs-180122.RData") #use below for statistical analysis
```

### Apply Divnet
Focus on genus and species level.
```{r,eval=FALSE}
###
gen_divnet <- divnet(tax_glom(physeq_insitu, taxrank = "Genus"), base = 3)
gen_divnet_label <- divnet(tax_glom(physeq_insitu, taxrank = "Genus"), X = "SAMPLELABEL", base = 3)
# Vent vs plume vs background
gen_divnet_TYPE <- divnet(tax_glom(physeq_insitu, taxrank = "Genus"), X = "SAMPLETYPE", base = 3)
# location and vent vs plume vs background
gen_divnet_TYPE_SITE <- divnet(tax_glom(physeq_insitu, taxrank = "Genus"), X = "TYPE_SITE", base = 3)
save(gen_divnet, gen_divnet_label, gen_divnet_TYPE, gen_divnet_TYPE_SITE, file = "GENUS.Rdata")

###
## Analyzed at the ASV level:
spp_divnet <- divnet(tax_glom(physeq_insitu, taxrank = "Species"), base = 3)

spp_divnet_label <- divnet(tax_glom(physeq_insitu, taxrank = "Species"), X = "SAMPLELABEL", base = 3)
# Vent vs plume vs background
spp_divnet_TYPE <- divnet(tax_glom(physeq_insitu, taxrank = "Species"), X = "SAMPLETYPE", base = 3)
# location and vent vs plume vs background
spp_divnet_TYPE_SITE <- divnet(tax_glom(physeq_insitu, taxrank = "Species"), X = "TYPE_SITE", base = 3)
save(spp_divnet, spp_divnet_label, spp_divnet_TYPE, spp_divnet_TYPE_SITE, file = "SPECIES.Rdata")
```

## Extract DivNet output and visualize
Above run on HPC and RData files save so we can look at various levels
of species richness.

Function to extract shannon and simpson data from each divnet output.

```{r extract shannon and simpson from divnet output}
fxn_extract_divet <- function(df){
  df$shannon %>% summary %>% 
  pivot_longer(cols = starts_with("estimate"), names_to = "ESTIMATE-shannon", values_to = "Shannon") %>%
  pivot_longer(cols = starts_with("error"), names_to = "ERROR-shannon", values_to = "Shannon-error") %>%
  pivot_longer(cols = starts_with("lower"), names_to = "LOWER-shannon", values_to = "Shannon-lower") %>%
  pivot_longer(cols = starts_with("upper"), names_to = "UPPER-shannon", values_to = "Shannon-upper") %>%
  left_join(df$simpson %>% summary %>%
      pivot_longer(cols = starts_with("estimate"), names_to = "ESTIMATE-simpson", values_to = "Simpson") %>%
      pivot_longer(cols = starts_with("error"), names_to = "ERROR-simpson", values_to = "Simpson-error") %>%
      pivot_longer(cols = starts_with("lower"), names_to = "LOWER-simpson", values_to = "Simpson-lower") %>%
      pivot_longer(cols = starts_with("upper"), names_to = "UPPER-simpson", values_to = "Simpson-upper"),
    by = c("sample_names" = "sample_names")) %>%
  left_join(metadata_insitu %>% rownames_to_column(var = "sample_names")) %>%
  select(-sample_names, -ends_with("-simpson"), -ends_with("-shannon"), -starts_with("model."), -starts_with("name.")) %>%
  distinct()
}
```

###Function to create plots

```{r}
### Set colors and symbols for dataset
sampletype_order <- c("Background", "Plume", "Vent")
site_symbol<- c(21, 23, 24, 22)
site_order <- c("Axial", "GordaRidge", "Piccard", "VonDamm")
site_color <- c("#fdbb84", "#31a354", "#ef3b2c", "#02818a")

###

plot_sampletype <- function(df){
  df$SITE_ORDER <- factor(df$SITE, levels = site_order)
  df$SAMPLETYPE_ORDER <- factor(df$SAMPLETYPE, levels = sampletype_order)
  names(site_symbol) <- site_order
  plot_grid(df %>%
  ggplot(aes(x = SAMPLETYPE_ORDER, y = Shannon, group = SAMPLETYPE_ORDER)) +
    geom_violin(color = "#525252", fill = "#525252", alpha = 0.2, width = 0.6, draw_quantiles = c(0.25, 0.5, 0.75)) +
    geom_jitter(color = "white", fill = "black", width = 0.2, size = 2, aes(shape = SITE_ORDER)) +
    scale_shape_manual(values = site_symbol) +
  theme_linedraw() +
  theme(axis.text.y = element_text(size = 14),
        axis.text.x = element_blank(),
        strip.background = element_blank(),
        strip.text = element_text(color = "black"),
        legend.position = "none",
        axis.ticks.x = element_blank()) +
  labs(x = "", y = "Shannon"),
  df %>%
    ggplot(aes(x = SAMPLETYPE_ORDER, y = Simpson, group = SAMPLETYPE_ORDER)) +
    geom_violin(color = "#525252", fill = "#525252", alpha = 0.1, width = 0.6, draw_quantiles = c(0.25, 0.5, 0.75)) +
    geom_jitter(color = "white", fill = "black", width = 0.2, size = 2.5, aes(shape = SITE_ORDER)) +
    scale_shape_manual(values = site_symbol) +
  theme_linedraw() +
  theme(axis.text.x = element_text(vjust = 1, hjust = 0.5, size = 14),
        axis.text = element_text(size = 14),
        strip.background = element_blank(),
        strip.text = element_blank(),
        legend.title = element_blank(),
        legend.position = "none") +
  labs(x = "", y = "Simpson"),
  ncol = 1, axis = c("lrt"), align = c("vh"))
}
```


### Plot Divnet results
Below commands use output from DivNet at different taxonomic levels to determine if pattern of Shannon index is consistent across taxonomic levels, which it appears to be. For final work, will use the species level (ASV level). 

```{r, fig.height=5, fig.width=3}
load("data-input/SPECIES.Rdata", verbose = T)
spp_alpha_18s <- fxn_extract_divet(spp_divnet)
spp_alpha_label <- fxn_extract_divet(spp_divnet_label)
spp_alpha_TYPE <- fxn_extract_divet(spp_divnet_TYPE)
spp_alpha_TYPE_SITE <- fxn_extract_divet(spp_divnet_TYPE_SITE)

plot_sampletype(spp_alpha_18s)

```

```{r, fig.height=4, fig.width=5}
spp_alpha_label$SITE_ORDER <- factor(spp_alpha_label$SITE, levels = site_order)
spp_alpha_label$SAMPLETYPE_ORDER <- factor(spp_alpha_label$SAMPLETYPE, levels = sampletype_order)
names(site_symbol) <- site_order
sampletype_symbol<- c(21, 23, 24)
# names(sampletype_symbol) <- sampletype_order
# svg("../../../Manuscripts_presentations_reviews/deepsea-microeuk-ME/vectorgraphic-figures/fig4b-shannon.svg", h = 4, w = 5)
spp_alpha_label %>%
  ggplot(aes(x = SAMPLETYPE_ORDER, y = Shannon, group = SAMPLETYPE_ORDER)) +
    geom_violin(color = "#525252", fill = "#525252", alpha = 0.2, width = 0.6, draw_quantiles = c(0.25, 0.5, 0.75)) +
    geom_jitter(color = "white", fill = "black", width = 0.2, size = 2, aes(shape = SITE_ORDER)) +
    # scale_fill_manual(values = site_color) +
    scale_shape_manual(values = site_symbol) +
  theme_linedraw() +
  theme(axis.text.y = element_text(size = 12, color = "black", face = "bold"),
        axis.text.x = element_text(size = 12, color = "black", face = "bold"),
        axis.title = element_text(size = 12, color = "black", face = "bold"),
        strip.background = element_blank(),
        strip.text = element_text(color = "black", face = "bold"),
        legend.position = "none",
        # legend.position = "right",
        legend.title = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(x = "", y = "Shannon")
# dev.off()
```
What vents or samples have the highest Shannon index?
```{r}
# View(spp_alpha_label)
```


Visualize with estimator variance included.
```{r, fig.height=5, fig.width=6}
# head(spp_alpha_18s)
# View(spp_alpha_label)
# View(spp_alpha_TYPE_SITE)
##
plot_grid(spp_alpha_label %>% 
  select(VENT, SITE, SAMPLETYPE, starts_with("Shannon"), starts_with("Simpson")) %>% 
  distinct() %>% 
  ggplot(aes(x = VENT, y = Shannon, fill = SITE, shape = SAMPLETYPE)) +
  geom_errorbar(aes(ymin = `Shannon-lower`, ymax = `Shannon-upper`), color = "#525252", width = 0.2) +
  geom_point(color = "#525252", size = 2, aes(fill = SITE, shape = SAMPLETYPE)) +
  scale_shape_manual(values = sampletype_symbol) +
  scale_fill_manual(values = site_color) +
  # coord_flip() +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 0),
        panel.background = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_text(color = "black", angle = 0, vjust = 0.5, hjust = 0.5),
        legend.position = "none",
        panel.grid = element_blank(),
        axis.ticks.y = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line()) +
  facet_grid(SITE + SAMPLETYPE ~ ., scale = "free", space = "free") +
    labs(x = "", y = "Shannon"),
#
spp_alpha_label %>% 
  select(VENT, SITE, SAMPLETYPE, starts_with("Shannon"), starts_with("Simpson")) %>% 
  distinct() %>% 
  ggplot(aes(x = VENT, y = Simpson, fill = SITE, shape = SAMPLETYPE)) +
  geom_errorbar(aes(ymin = `Simpson-lower`, ymax = `Simpson-upper`), color = "#525252", width = 0.2) +
  geom_point(color = "#525252", size = 2, aes(fill = SITE, shape = SAMPLETYPE)) +
  scale_shape_manual(values = sampletype_symbol) +
  scale_fill_manual(values = site_color) +
  # coord_flip() +
  theme_linedraw() +
  theme(axis.text.y = element_blank(),
        panel.background = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_text(color = "black", angle = 0, vjust = 0.5, hjust = 0.5),
        legend.position = "none",
        panel.grid = element_blank(),
        axis.ticks.y = element_blank(),
        panel.border = element_blank(),
        axis.line.x = element_line()) +
  facet_grid(SITE + SAMPLETYPE ~ ., scale = "free", space = "free") +
    labs(x = "", y = "Simpson"),
ncol = 2, rel_widths = c(1,0.7))


shannon_divnet <- spp_alpha_label %>% 
  select(VENT, SITE, SAMPLETYPE, starts_with("Shannon"), starts_with("Simpson")) %>% 
  distinct() %>% 
  ggplot(aes(x = VENT, y = Shannon, fill = SITE, shape = SAMPLETYPE)) +
  geom_errorbar(aes(ymin = `Shannon-lower`, ymax = `Shannon-upper`), color = "#525252", width = 0.2) +
  geom_point(color = "#525252", size = 2, aes(fill = SITE, shape = SAMPLETYPE)) +
  scale_shape_manual(values = sampletype_symbol) +
  scale_fill_manual(values = site_color) +
  # coord_flip() +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        panel.background = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_text(color = "black", angle = 0, vjust = 0.5, hjust = 0.5),
        legend.position = "none",
        panel.grid = element_blank(),
        axis.ticks.y = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line()) +
  facet_grid(. ~ SITE + SAMPLETYPE, scale = "free", space = "free") +
    labs(x = "", y = "Shannon")
```

Figure S6.
```{r, fig.height=4, fig.width=9}
# svg("../../../Manuscripts_presentations_reviews/deepsea-microeuk-ME/vectorgraphic-figures/figS6.svg", h=4, w = 9)

shannon_divnet
# dev.off()
```

# Network analysis

Import data frames below and format as phyloseq. Subsample to select ASVs that appear in more than 1 sample and have a total of 100 or more sequences. This leaves 2577 total ASVs across 43 samples.

> Run below commands on HPC

```{r, eval=FALSE}
load("asv-tables-processed-18102021.RData", verbose = T)
```
```{r, eval=FALSE}
library(phyloseq)
# library(SpiecEasi)
```

```{r Input to matrix for phyloseq object, eval=FALSE}
# Select eukaryotes only and create wide format dataframe
insitu_wide_nosingle <- asv_insitu_qc %>%
  filter(Domain == "Eukaryota") %>%
  filter(!grepl("_Plume001_", SAMPLE)) %>% #removing "near vent background", not relevant in other data sets
  select(FeatureID, Taxon, SAMPLE, value) %>%
  pivot_wider(names_from = SAMPLE, values_from = value, values_fill = 0) %>%
  mutate(PREVALENCE = rowSums(select_if(., is.numeric) > 0),
         SEQ_TOTAL = rowSums(select_if(., is.numeric))) %>%
  filter(PREVALENCE >= 1) %>% 
  filter(SEQ_TOTAL >= 100)
#
# ASVs had to appear in more than 1 sample and have at least 100 sequences
length(unique(insitu_wide_nosingle$FeatureID))
length(unique(asv_insitu_qc$FeatureID))

# head(insitu_wide)
insitu_samples <- as.character(colnames(insitu_wide_nosingle %>% select(-Taxon, -FeatureID)))


# make matrices for phyloseq
insitu_tax_matrix <- insitu_wide_nosingle %>%
  select(FeatureID, Taxon) %>%
  separate(Taxon, c("Domain", "Supergroup",
                  "Phylum", "Class", "Order",
                  "Family", "Genus", "Species"), sep = ";") %>%
  column_to_rownames(var = "FeatureID") %>%
  as.matrix

insitu_asv_matrix <- insitu_wide_nosingle %>%
  select(-Taxon) %>%
  column_to_rownames(var = "FeatureID") %>%
  as.matrix


# Align row names for each matrix
rownames(insitu_tax_matrix) <- row.names(insitu_asv_matrix)

metadata_insitu <- metadata %>%
  filter(SAMPLE %in% insitu_samples) %>% # from reformatting df above
  select(SAMPLE, VENT, SITE, SAMPLETYPE, YEAR) %>%
  unite(SAMPLELABEL, VENT, SITE, SAMPLETYPE, YEAR, sep = "_", remove = FALSE) %>%
  unite(TYPE_SITE, SITE, SAMPLETYPE, sep = "_", remove = FALSE)

rownames(metadata_insitu) <- metadata_insitu$SAMPLE

# Import asv and tax matrices
ASV = otu_table(insitu_asv_matrix, taxa_are_rows = TRUE)
TAX = tax_table(insitu_tax_matrix)

phylo_obj <- phyloseq(ASV, TAX)

# Import metadata as sample data in phyloseq
samplenames <- sample_data(metadata_insitu)

## Check
physeq_insitu
```

Run SPIEC-EASI with phyloseq object.
```{r, eval=FALSE}
# Run spiec easi with glasso
pargs2 <- list(rep.num = 50, seed = 10010, ncores = 10)
spec_glasso_microeuk <- spiec.easi(physeq_insitu, method = 'glasso', lambda.min.ratio=1e-2, nlambda=20,pulsar.params=pargs2)

# save(spec_glasso_microeuk, file = "spiec-easi-output-03-12-21.RData")
```

## Results from 12-8-2021
Isolate ASV-ASV pairs of interest
```{r, eval=FALSE}
# load("spiec-easi-output-03-12-21.RData", verbose = T) # almost 5GB file!

getStability(spec_glasso_microeuk) # Target == 0.05
# [1] 0.03827056
sum(getRefit(spec_glasso_microeuk))/2
# [1] 45904.5

# spec_glasso_microeuk
# Pulsar-selected refit of sparseiCov 
# Path length: 20 
# Graph dim:   2577 
# Criterion:
#   stars... sparsity 0.0138
```

Extract weighted matrix
```{r, eval=FALSE}
# se_beta <- as.matrix(symBeta(getOptBeta(spec_glasso_microeuk)))
# df_beta <- as.data.frame(se_beta)

# Extract weight information
glasso_weight  <- cov2cor(as.matrix(getOptCov(spec_glasso_microeuk)))

colnames(glasso_weight) <- rownames(glasso_weight)# <- colnames(Networ_taxa_DF_WideMat) #here i may be able to give the verticies the taxa names if i feed it the vector of names from the levels.

weighted_adj_mat <- glasso_weight*getRefit(spec_glasso_microeuk)
df_weighted <- as.data.frame(as.matrix(weighted_adj_mat))
# Assign column and row names - from original glasso output matrix data
colnames(df_weighted) <- colnames(spec_glasso_microeuk$est$data)
row.names(df_weighted) <- colnames(spec_glasso_microeuk$est$data)
```

Work with weighted dataframe
```{r, eval=FALSE}
key <- insitu_asv_wClass %>%
  select(FeatureID, Taxon, Domain:Species, CLASS, SITE_CLASS) %>% 
  distinct()
# head(key)

df_spieceasi <- df_weighted %>% 
 rownames_to_column(var = "sideA") %>% 
  pivot_longer(cols = -sideA, names_to = "sideB") %>% 
  left_join(key, by = c(sideA = "FeatureID")) %>% 
  left_join(key, by = c(sideB = "FeatureID"), suffix = c("_sideA", "_sideB")) %>% 
  distinct()

# 6640929 total interactions
## 6 million edged?
df_spieceasi_filtered <- df_spieceasi %>% 
  filter(abs(value) > 0.01) %>% 
  mutate(Interaction = case_when(
    value < 0 ~ "negative",
    value > 0 ~ "positive"
  ))
# Leaves 91,288 interactions
## Interaction type
# negative positive 
    # 3363    87925 

save(df_spieceasi_filtered, file = "filtered-spieceasi-result-08122021.RData")  
```

## Bring results locally
```{r}
load("data-input/filtered-spieceasi-result-08122021.RData", verbose = TRUE)
```
> H0: the majority of protist-protist pairs will reveal host-parasite interactions, and then predator-prey.

Questions to ask regarding the SPIEC EASI results.
- What is the overall taxonomic composition of negative and positive co-occurring ASVs?
- What percentage of putative interactions include likely parasitic protists?

Format, get summary stats from network analyss.
```{r, fig.width=18, fig.height=7}
# head(df_spieceasi_filtered)

spieceasi_rm_reps <- df_spieceasi_filtered %>% 
  mutate(TMP_ASV_REP = purrr::map2_chr(sideA, sideB, ~toString(sort(c(.x, .y))))) %>%
  select(TMP_ASV_REP, value, Interaction) %>% 
    group_by(TMP_ASV_REP, value, Interaction) %>% 
    distinct() %>% 
  ungroup() %>% 
  separate(TMP_ASV_REP, c("sideA", "sideB"), sep = ", ") %>% 
  left_join((select(df_spieceasi_filtered, ends_with("sideA")) %>% distinct())) %>% 
  left_join((select(df_spieceasi_filtered, ends_with("sideB")) %>% distinct()))

head(spieceasi_rm_reps)
```
Look at highest percentage of ASV-ASV pairs by various categories.
```{r}
# Get stats
totaloccur <- dim(spieceasi_rm_reps)[1]

spieceasi_rm_reps %>% 
  unite(CLASS_joined, CLASS_sideA, CLASS_sideB, sep = "-") %>%
  unite(SITE_CLASS_joined, SITE_CLASS_sideA, SITE_CLASS_sideB, sep = "-") %>%
  unite(PHYLUM_joined, Phylum_sideA, Phylum_sideB, remove = FALSE, sep = "-") %>%
  group_by(CLASS_joined, Interaction) %>%
  summarise(COUNT = n(),
            PERC = 100*(COUNT/totaloccur)) %>% 
  arrange(desc(PERC))
```
> ASV-ASV pairs were primarily among ASVs classified as 'Vent only' - both ASVs were resident. 36%

> Secondly, 15% of the ASV pairs were from ASVs specifically found in all sample types.

> Again, mostly all positive. The highest occurence of negative interactions were between Vent only- Vent, plume, and background. 

```{r}
asv_asv <- c("Vent only-Vent only", "Vent, plume, & background-Vent, plume, & background")

# Get stats
totaloccur <- dim(spieceasi_rm_reps)[1]

spieceasi_rm_reps %>% 
  unite(CLASS_joined, CLASS_sideA, CLASS_sideB, sep = "-") %>%
  unite(SITE_CLASS_joined, SITE_CLASS_sideA, SITE_CLASS_sideB, sep = "-") %>%
  unite(PHYLUM_joined, Phylum_sideA, Phylum_sideB, remove = FALSE, sep = "-") %>%
  filter(CLASS_joined %in% asv_asv) %>% 
  group_by(CLASS_joined) %>% 
    mutate(TOTALCOUNT = n()) %>% 
  group_by(CLASS_joined, PHYLUM_joined, Interaction, TOTALCOUNT) %>%
  summarise(COUNT = n(),
            PERC = 100*(COUNT/TOTALCOUNT)) %>%
  ungroup() %>% 
  arrange(desc(PERC)) %>% 
  distinct() %>% 
  select(CLASS_joined, PHYLUM_joined, PERC, Interaction) %>% 
  pivot_wider(names_from = CLASS_joined, values_from = PERC)
```
> Within 'Vent only-Vent only' ASV pairs, almost 15% were among ciliates (ciliate-ciliate), while for the 'Vent, plume, & backgroun' ASV pairs, almost 13% were between dinoflagellates.

Isolate the ASV-ASV pairs that appeared most frequently. These include ciliates, dinoflagellates, and radiolaria within the resident and cosmopolitan (the latter includes ASVs that appears at least once ALL sample types).
```{r}
rad_dino <- c("Dinoflagellata-Radiolaria", "Radiolaria-Dinoflagellata", "Radiolaria-Radiolaria")

ciliate_ciliate <- spieceasi_rm_reps %>% 
  unite(CLASS_joined, CLASS_sideA, CLASS_sideB, sep = "-") %>%
  unite(SITE_CLASS_joined, SITE_CLASS_sideA, SITE_CLASS_sideB, sep = "-") %>%
  unite(PHYLUM_joined, Phylum_sideA, Phylum_sideB, remove = FALSE, sep = "-") %>%
  filter(CLASS_joined %in% asv_asv) %>%
  filter(PHYLUM_joined == "Ciliophora-Ciliophora")

dino_dino <- spieceasi_rm_reps %>% 
  unite(CLASS_joined, CLASS_sideA, CLASS_sideB, sep = "-") %>%
  unite(SITE_CLASS_joined, SITE_CLASS_sideA, SITE_CLASS_sideB, sep = "-") %>%
  unite(PHYLUM_joined, Phylum_sideA, Phylum_sideB, remove = FALSE, sep = "-") %>%
  filter(CLASS_joined %in% asv_asv) %>%
  filter(PHYLUM_joined == "Dinoflagellata-Dinoflagellata")

rad_dino <- spieceasi_rm_reps %>% 
  unite(CLASS_joined, CLASS_sideA, CLASS_sideB, sep = "-") %>%
  unite(SITE_CLASS_joined, SITE_CLASS_sideA, SITE_CLASS_sideB, sep = "-") %>%
  unite(PHYLUM_joined, Phylum_sideA, Phylum_sideB, remove = FALSE, sep = "-") %>%
  filter(CLASS_joined %in% asv_asv) %>%
  filter(PHYLUM_joined %in% rad_dino)
```

```{r}
total <- dim(spieceasi_rm_reps)[1]
head(dino_dino)
(dim(dino_dino)[1]/total) * 100
(dim(ciliate_ciliate)[1]/total) * 100

by_phylum <- (spieceasi_rm_reps %>% unite(PHYLUM_joined, Phylum_sideA, Phylum_sideB, remove = FALSE, sep = "-"))

```


### Ciliate-ciliate
Isolate associations with ciliates
```{r, fig.height=15, fig.width=15}
# head(ciliate_ciliate)
# table(ciliate_ciliate$SITE_CLASS_joined)
# table(ciliate_ciliate
ciliate_ciliate %>% 
  filter(CLASS_joined == "Vent only-Vent only") %>% 
  # filter(!is.na(Phylum_sideA) | !is.na(Phylum_sideB)) %>%
  ggplot(aes(x = Taxon_sideA, y = Taxon_sideB, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient(low = "#bfd3e6", high = "#810f7c") +
  theme_linedraw() +
  facet_grid(Class_sideB +  Order_sideB ~ Class_sideA + Order_sideA, space = "free", scales = "free") +
  theme(axis.text = element_blank(),
      panel.grid.minor = element_blank(),
      strip.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5),
      strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
      strip.text = element_text(color = "black", face = "bold"),
      strip.background = element_blank())
```
Isolate functional traits from these groups of ciliates
```{r}
# order <- c("Plagiopylea", "Nassophorea", "Litostomatea")
# family <- c("Scuticociliatia", "Euplotia", "Suctoria", "Peritrichia")
# genus <- c("Strombidiidae")
```


Tile plot of vent-only (resident) ciliate-ciliate ASV co-occurrences.
```{r, fig.height=8, fig.width=10}
# head(ciliate_ciliate)
# hist(ciliate_ciliate$value)

ciliate_ciliate %>% 
       filter(value >= 0.4) %>% 
       select(starts_with(c("Class_", "Order_")), PHYLUM_joined) %>% 
       select(-CLASS_joined) %>% 
       # select(Taxon_sideA, Taxon_sideB, PHYLUM_joined) %>% 
       unite(SIDEA, Class_sideA, Order_sideA, sep = ";", remove = FALSE) %>%
        unite(SIDEB, Class_sideB, Order_sideB, sep = ";", remove = FALSE) %>% 
      group_by(SIDEA, SIDEB, Class_sideA, Class_sideB) %>% 
      summarise(COUNT = n()) %>% 
  ggplot(aes(x = SIDEA, y = SIDEB, fill = COUNT)) +
    geom_tile(color = "black") +
  scale_fill_gradient(low = "#bfd3e6", high = "#810f7c") +
  theme_linedraw() +
  facet_grid(Class_sideB ~ Class_sideA, space = "free", scales = "free") +
  theme(panel.grid.minor = element_blank(),
      strip.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5),
      strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
      strip.text = element_text(color = "black", face = "bold"),
      strip.background = element_blank(),
      axis.title = element_blank(),
      axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5))
      
```



### Dinoflagellate
```{r, fig.height=7, fig.width=8}
dino_dino %>% 
  filter(CLASS_joined == "Vent only-Vent only") %>% 
  # filter(!is.na(Phylum_sideA) | !is.na(Phylum_sideB)) %>%
  ggplot(aes(x = Taxon_sideA, y = Taxon_sideB, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient(low = "#bfd3e6", high = "#810f7c") +
  theme_linedraw() +
  facet_grid(Class_sideB +  Order_sideB ~ Class_sideA + Order_sideA, space = "free", scales = "free") +
  theme(axis.text = element_blank(),
      panel.grid.minor = element_blank(),
      strip.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5),
      strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
      strip.text = element_text(color = "black", face = "bold"),
      strip.background = element_blank())
```
Dinos and others
Get stats on Ciliate-other ASV pairs for a table
```{r}

```

### Retaria and dino

```{r, fig.height=12, fig.width=12}
rad_dino %>% 
  filter(CLASS_joined == "Vent only-Vent only") %>% 
  # filter(!is.na(Phylum_sideA) | !is.na(Phylum_sideB)) %>%
  ggplot(aes(x = Taxon_sideA, y = Taxon_sideB, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient(low = "#bfd3e6", high = "#810f7c") +
  theme_linedraw() +
  facet_grid(Class_sideB +  Order_sideB ~ Class_sideA + Order_sideA, space = "free", scales = "free") +
  theme(axis.text = element_blank(),
      panel.grid.minor = element_blank(),
      strip.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5),
      strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
      strip.text = element_text(color = "black", face = "bold"),
      strip.background = element_blank())
```

## Isolate most frequent co-occurrences
Information for table 3.
```{r}
head(spieceasi_rm_reps)
```

```{r}
freq <- spieceasi_rm_reps %>% 
  unite(CLASS_joined, CLASS_sideA, CLASS_sideB, sep = "-") %>%
  filter(!(is.na(Phylum_sideA)) | !is.na(Phylum_sideB)) %>% 
  unite(PHYLUM_joined, Phylum_sideA, Phylum_sideB, remove = FALSE, sep = "-") %>%
  group_by(CLASS_joined, PHYLUM_joined, Taxon_sideA, Taxon_sideB) %>% 
    summarise(FREQUENCY = n()) %>% 
  arrange(desc(FREQUENCY)) %>% 
  filter(FREQUENCY > 2)

# View(freq %>% filter(grepl("Cercozoa-", PHYLUM_joined)))
# View(freq %>% filter(grepl("Haptophyta-", PHYLUM_joined)))
# View(freq %>% filter(grepl("Opalozoa-", PHYLUM_joined) | grepl("Pseudofungi-", PHYLUM_joined)))
# View(freq %>% filter(grepl("Radiolaria-", PHYLUM_joined)))
# range(freq$FREQUENCY)
```


## Test influence of abiotic/biotic factors on community composition.

### Set up data and test for outliers
Import phyloseq-saved RObjects.
```{r}
# library(tidyverse); library(phyloseq); library(compositions); library(vegan)
load("phyloseq-objs-180122.RData", verbose = TRUE)
```


Isolate ASV count table using phyloseq commands and perform a CLR transformation.
```{r}
# Extract phyloseq object & CLR transform
count_table_clr <- compositions::clr(otu_table(physeq_insitu))
# sample_data(physeq_insitu)
?clr()
count_table_notrans <- otu_table(physeq_insitu)
```

Check all samples considered and estimate distance matrix (Euclidean)
```{r}
# transpose and make dist matrix
sample_data(physeq_insitu)

# CLR-transformed data, transpose and create euclidean dist matrix
dist_euc <- vegdist(t(count_table_clr), method = "euclidean")
# ?vegdist()
# ?betadisper

# dist_jac <- vegdist(t(count_table_notrans), method = "jaccard")
```

```{r}
# ?betadisper()
beta_SITE <- betadisper(dist_euc, sample_data(physeq_insitu)$SITE)
beta_SAMPLETYPE <- betadisper(dist_euc, sample_data(physeq_insitu)$SAMPLETYPE)
beta_TYPE_SITE <- betadisper(dist_euc, sample_data(physeq_insitu)$TYPE_SITE)
```

SITE
```{r}
anova(beta_SITE)
TukeyHSD(beta_SITE)
plot(beta_SITE)
# View(beta_SITE$vectors) # to confirm sample that is an outlier
```

Sampletype
```{r}
anova(beta_SAMPLETYPE)
TukeyHSD(beta_SAMPLETYPE)
plot(beta_SAMPLETYPE)
# View(beta_SAMPLETYPE$vectors)
```
Based on dispersion testing, it is clear that Boca (Axial, snowblower vent) is an outlier. Secondary to Boca, however are the plume and CTD (Background samples).

#### Remove outlier
Repeat dispersion test with removed samples, Boca and background.

```{r}
count_table_clr_noboca <- compositions::clr(otu_table(subset_samples(physeq_insitu, VENT != "Boca")))

count_table_clr_nobocaBSW <- compositions::clr(otu_table(subset_samples(physeq_insitu, (VENT != "Boca" & SAMPLETYPE == "Vent"))))
# names(count_table_clr_nobocaBSW)

# CLR-transformed data, transpose and create euclidean dist matrix
dist_euc_noboca <- vegdist(t(count_table_clr_noboca), method = "euclidean")

dist_euc_nobocaBSW <- vegdist(t(count_table_clr_nobocaBSW), method = "euclidean")
# ?vegdist()
```

Re-check dispersion without Boca sample.
```{r}
beta_SITE <- betadisper(dist_euc_noboca, sample_data(subset_samples(physeq_insitu, VENT != "Boca"))$SITE)
anova(beta_SITE)
TukeyHSD(beta_SITE)
plot(beta_SITE)
# View(beta_SITE$vectors) # to confirm samples that are outliers
```

Check dispersion without Boca or background samples.
```{r}
beta_SITE <- betadisper(dist_euc_nobocaBSW, sample_data(subset_samples(physeq_insitu, (VENT != "Boca" & SAMPLETYPE == "Vent")))$SITE)
anova(beta_SITE)
TukeyHSD(beta_SITE)
plot(beta_SITE)
# View(beta_SITE$vectors) # to confirm sample that is an outlier
```

## Distance-based redundancy analysis

Legendre and Anderson (1999), RDA that looks for linear relationships on dissimilarity matrices. Constrained ordination with non-Euclidean distance measures. (Legendre P, Anderson MJ (1999) Distance-based redundancy analysis: testing multispecies responses in multifactorial ecological experiments. Ecol Monogr. 69:1–24.)

Below ```dbrda()``` statement inputs all samples as a distance matrix (Euclidean), with the RH side variables as numeric data following the ~. 

### Function for DBRDA analysis

With various outliers depending on the question posed, below function takes subseted data and runs ```dbrda()``` analysis.

dbrda() analysis input different subsets of phyloseq object, perform Jaccard dissimilarity.

```{r}
dbrda_subset <- function(subset_df){
  # category <- enquo(category)
  # subset_df <- subset_samples(physeq_insitu, category %in% selection)
  clr_out <- compositions::clr(otu_table(subset_df))
  euc_out <- vegdist(t(clr_out), method = "euclidean") #CLR with Euclidean is aitchison
  # jac_out <- vegdist(t(otu_table(subset_df)), method = "jaccard")
  # Params
  # ?vegdist
  temp <- sample_data(subset_df)$TEMP
  ph <- sample_data(subset_df)$pH
  persea <- sample_data(subset_df)$PercSeawater
  mg <- sample_data(subset_df)$Mg
  h2 <- sample_data(subset_df)$H2
  h2s <- sample_data(subset_df)$H2S
  ch4 <- sample_data(subset_df)$CH4
  prok <- sample_data(subset_df)$ProkConc
  depth <- sample_data(subset_df)$DEPTH
  # DBRDA
   rda_out <- dbrda(euc_out ~ temp + persea + prok + h2s + mg + h2 + ch4 + depth + ph, dist = vegdist, na.action = na.omit)
  # Overall significance
  # anova(rda_out)
  # Print results by margin
  anova(rda_out, by = 'margin')
  # Plot
  # xy <- anova(rda_out, by = 'axis')
  # x <- round(xy$Variance[1], 1); y <- round(xy$Variance[2], 1)
  # plot(rda_out, xlab = x, ylab = y)
}

```
Usage of function, input phyloseq object with all data, it will be transformed (CLR) and dist matrix will be estimates (Euclidean). Parameters from the metadata will be formatted and dbrda() will run on all. 

> CLR transformed data with Euclidean distance estimate is Aitchison's

## Test sample subsets by sample type and taxa

Removed Boca outlier and isolate vent-only sites, test for significant parameters.
```{r}
dbrda_subset(subset_samples(physeq_insitu, (VENT != "Boca" & SAMPLETYPE == "Vent")))
```

Isolate subsets of data frame by ASV or taxonomic ID.
```{r}
noNAs <- filter(insitu_asv_wClass, 
                (Supergroup != "Bacteria_X" | 
                   !is.na(Supergroup) | 
                   Supergroup != "Terrabacteria" | 
                   Supergroup != "Archaea_X" | 
                   Supergroup != "Opisthokonta"))
list_of_noNAs <- as.character(noNAs$FeatureID)
# length(list_of_noNAs)
noNAs_phy <- prune_taxa(list_of_noNAs, physeq_insitu)

###
cosmo <- c("Vent, plume, & background", "Vent & background", "Vent & plume", "Plume & background")

res <- filter(insitu_asv_wClass, CLASS == "Vent only")
cos <- filter(insitu_asv_wClass, CLASS %in% cosmo)

list_of_resident <- as.character(unique(res$FeatureID))
list_of_cos <- as.character(unique(cos$FeatureID))
# length(list_of_resident); length(list_of_cos)
cosmo_only <- prune_taxa(list_of_cos, physeq_insitu)
res_vent <- prune_taxa(list_of_resident, physeq_insitu)

###
# sample_data(insitu_asv_wClass)
alvs <- filter(insitu_asv_wClass, Supergroup == "Alveolata")
list_of_alvs <- as.character(alvs$FeatureID)
alvs_phy <- prune_taxa(list_of_alvs, physeq_insitu)

##
# sample_data(insitu_asv_wClass)
dino_syn <- filter(insitu_asv_wClass, (Phylum == "Dinoflagellata" | Class == "Syndiniales"))
list_of_dino_syn <- as.character(dino_syn$FeatureID)
dino_syn_phy <- prune_taxa(list_of_dino_syn, physeq_insitu)
##
# sample_data(insitu_asv_wClass)
cili <- filter(insitu_asv_wClass, Phylum == "Ciliophora")
list_of_cili <- as.character(cili$FeatureID)
cili_phy <- prune_taxa(list_of_cili, physeq_insitu)

##
# sample_data(insitu_asv_wClass)
rhiz <- filter(insitu_asv_wClass, (Supergroup == "Rhizaria"))
list_of_rhiz <- as.character(rhiz$FeatureID)
rhiz_phy <- prune_taxa(list_of_rhiz, physeq_insitu)

##
# sample_data(insitu_asv_wClass)
stram <- filter(insitu_asv_wClass, (Supergroup == "Stramenopiles"))
list_of_stram <- as.character(stram$FeatureID)
stram_phy <- prune_taxa(list_of_stram, physeq_insitu)
```


### Table 2 tests
```{r}
set.seed(1245)
```

Test with data set, vents only. Populate Table 2.
```{r}
# All samples, without Boca as an outlier

dbrda_output <- function(df, inputdata){
  allsamples <- dbrda_subset(subset_samples(df, (VENT != "Boca")))
    allsamples_df <- data.frame(variables = row.names(allsamples), sig = allsamples$`Pr(>F)`, samples = "All samples") %>% drop_na()
    
    # vent samples only, Boca removed
    ventonly <- dbrda_subset(subset_samples(df, (SAMPLETYPE == "Vent" & VENT != "Boca")))
    ventonly_df <- data.frame(variables = row.names(ventonly), sig = ventonly$`Pr(>F)`, samples = "Vent only") %>% drop_na()
    
    # MCR: Piccard and Von Damm, vents only
    mcr <- dbrda_subset(subset_samples(df, (SITE == "VonDamm" | SITE == "Piccard" | SAMPLETYPE == "Vent")))
    mcr_df <- data.frame(variables = row.names(mcr), sig = mcr$`Pr(>F)`, samples = "MCR") %>% drop_na()
    
    # GR and Axial
    NEpac <- dbrda_subset(subset_samples(df, (SITE == "GordaRidge" | SITE == "Axial" & VENT != "Boca" | SAMPLETYPE == "Vent")))
    NEpac_df <- data.frame(variables = row.names(NEpac), sig = NEpac$`Pr(>F)`, samples = "NE Pacific") %>% drop_na()
    
    # Von Damm and then Piccard only
    vd <- dbrda_subset(subset_samples(df, (SITE == "VonDamm" | SAMPLETYPE == "Vent")))
    vd_df <- data.frame(variables = row.names(vd), sig = vd$`Pr(>F)`, samples = "Von Damm") %>% drop_na()
    
    picc <- dbrda_subset(subset_samples(df, (SITE == "Piccard" | SAMPLETYPE == "Vent")))
    picc_df <- data.frame(variables = row.names(picc), sig = picc$`Pr(>F)`, samples = "Piccard") %>% drop_na()
    
    # Axial and then Gorda only
    axial <- dbrda_subset(subset_samples(df, (VENT != "Boca" & SITE == "Axial" | SAMPLETYPE == "Vent")))
    axial_df <- data.frame(variables = row.names(axial), sig = axial$`Pr(>F)`, samples = "Axial Seamount") %>% drop_na()
    
    gr <- dbrda_subset(subset_samples(df, (SITE == "GordaRidge" | SAMPLETYPE == "Vent"))); gr
    gr_df <- data.frame(variables = row.names(gr), sig = gr$`Pr(>F)`, samples = "Gorda Ridge") %>% drop_na()
    
    table2_output <- bind_rows(
      allsamples_df, ventonly_df, mcr_df, NEpac_df, vd_df, picc_df, axial_df, gr_df) %>% 
      add_column(Taxa = inputdata)
    return(table2_output)
}

```


```{r}
table2_stats <- bind_rows(
  dbrda_output(physeq_insitu, "All taxa"),
  dbrda_output(noNAs_phy, "No NAs, no metazoa"),
  dbrda_output(res_vent, "Resident only"),
  dbrda_output(cosmo_only, "Cosmopolitan only"),
  dbrda_output(alvs_phy, "Alveolates (all)"),
  dbrda_output(dino_syn_phy, "Dinoflagelaltes & Syndiniales"),
  dbrda_output(cili_phy, "Ciliates"),
  dbrda_output(rhiz_phy, "Rhizaria"),
  dbrda_output(stram_phy, "Stramenopiles"),
)
# View(table2_stats)
# head(table2_stats)
# unique(table2_stats$Taxa)
```

```{r, fig.height=4, fig.width=14}
# range(table2_output$sig)
reds <- c("#f7f7f7","#fee5d9", "#fcae91", "#fb6a4a", "#a50f15")

# Factor variables
# rm <- c("Von Damm", "Piccard", "Axial Seamount", "Gorda Ridge")

table2_stats$SAMPLES <- factor(table2_stats$samples, levels = c("All samples", "Vent only", "MCR", "NE Pacific", "Von Damm", "Piccard", "Axial Seamount", "Gorda Ridge"), labels = c("All samples", "Vent only", "Mid-Cayman Rise (vents)", "NE Pacific (vents)", "Von Damm", "Piccard", "Axial Seamount", "Gorda Ridge"))

order_taxa <- c("All taxa", "No NAs, no metazoa", "Resident only", "Cosmopolitan only", "Alveolates (all)", "Dinoflagelaltes & Syndiniales", "Ciliates", "Rhizaria", "Stramenopiles")
table2_stats$TAXA <- factor(table2_stats$Taxa, levels = rev(order_taxa))

table2_stats$VARIABLES <- factor(table2_stats$variables, 
                                 levels = c("depth", "temp", "prok", "ph", "persea", "mg", "h2", "h2s", "ch4"), 
                                 labels = c("Depth (m)", "Temperature (C)", "Cells ml^-1", "pH", "Seawater %", "Mg", "H2", "H2S", "CH4"))

# svg("dbrda-outcome-heatmap.svg", h = 4, w = 14)
table2_stats %>% 
  # filter(!(samples %in% rm)) %>% 
  ggplot(aes(x = VARIABLES, y = TAXA, fill = sig)) +
    geom_tile(stat = "identity", color = "white") +
  facet_grid(. ~ SAMPLES) +
  scale_fill_stepsn(
    limits = c(0,0.25),
    breaks = c(0.01, 0.05, 0.1),
    show.limits = TRUE,
    colors = rev(reds)) +
  theme_bw() +
  scale_x_discrete(position = "top") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0, hjust = 0, color = "black"),
        strip.background = element_blank(),
        panel.border = element_blank(),
        axis.text.y = element_text(color = "black"),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        strip.placement = "outside") +
  labs(x = "Parameters by dataset", y = "Taxa")
# dev.off()
```
# Profiles by individual vent site

```{r}
axial <- c("Axial")
mcr <- c("VonDamm", "Piccard")
gr <- c("GordaRidge")
all <- c("Axial", "VonDamm", "Piccard", "GordaRidge")
```


## Bar plot function

Create a bar plot showing the relative sequence abundance of 18S results
to the Supergroup and Phylum level. Function averages across replicates
and then sums to the phylum and supergroup level. Bar plot shows the
relative sequence abundance.

```{r Barplot function to be used below, tidy=FALSE, fig.height=12, fig.width=9}
make_bar_relabun <- function(df, selection){
  df_out <- df %>% 
  filter(SITE %in% selection) %>% 
  filter(Domain == "Eukaryota") %>%
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  group_by(FeatureID, Taxon, Domain, Supergroup, Phylum, Class, Order, Family, Genus, Species,
           VENT, SITE, SAMPLETYPE, YEAR, DATASET) %>% 
    summarise(SEQ_AVG_REP = mean(value)) %>% 
  ungroup()
  supergroup <- df_out %>% 
  group_by(SITE, SAMPLETYPE, VENT, YEAR, Supergroup) %>% 
    summarise(SEQ_SUM = sum(SEQ_AVG_REP)) %>% 
  ggplot(aes(x = VENT, y = SEQ_SUM, fill = Supergroup)) +
    geom_bar(stat = "identity", position = "fill", color = "black", width = 0.9) +
    facet_grid(. ~ SITE +YEAR + SAMPLETYPE, scale = "free", space = "free") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        strip.background = element_blank(), strip.text = element_text(color = "black"),
        legend.position = "right") +
  scale_y_continuous(expand = c(0,0)) +
  # scale_fill_brewer(palette = "Set2") +
  scale_fill_manual(values = c("#fa9fb5", "#c51b8a", "#67000d", "#ef3b2c", "#ffffcc", "#feb24c", "#c7e9b4", "#1d91c0", "#253494", "#9e9ac8", "#238b45", "#54278f", "#bdbdbd", "#252525", "#fa9fb5", "#c51b8a", "#67000d", "#ef3b2c", "#ffffcc", "#feb24c", "#c7e9b4", "#1d91c0", "#253494", "#9e9ac8", "#238b45", "#54278f", "#bdbdbd", "#252525")) +
  labs(x = "", y = "Relative abundance")
  
  phylum <- df_out %>% 
    unite(SupergroupPhylum, Supergroup, Phylum, sep = "-") %>% 
  group_by(SITE, SAMPLETYPE, VENT, YEAR, SupergroupPhylum) %>% 
    summarise(SEQ_SUM = sum(SEQ_AVG_REP)) %>% 
  ggplot(aes(x = VENT, y = SEQ_SUM, fill = SupergroupPhylum)) +
    geom_bar(stat = "identity", position = "fill", color = "black", width = 0.9) +
    facet_grid(. ~ SITE +YEAR + SAMPLETYPE, scale = "free", space = "free") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        strip.background = element_blank(), strip.text = element_text(color = "black"),
        legend.position = "right") +
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_manual(values = c("#f1eef6", "#d7b5d8", "#df65b0", "#ce1256", "#fc9272", "#ef3b2c", 
    "#800026", "#fff7bc", "#fec44f", "#d95f0e", "#a63603", "#74c476", "#238b45", 
    "#00441b", "#7fcdbb", "#084081", "#c6dbef", "#2b8cbe", "#016c59", "#bcbddc", 
    "#807dba", "#54278f", "#bdbdbd", "black", "white", "#969696", "#525252", "#f1eef6", "#d7b5d8", "#df65b0", "#ce1256", "#fc9272", "#ef3b2c", 
    "#800026", "#fff7bc", "#fec44f", "#d95f0e", "#a63603", "#74c476", "#238b45", 
    "#00441b", "#7fcdbb", "#084081", "#c6dbef", "#2b8cbe", "#016c59", "#bcbddc", 
    "#807dba", "#54278f", "#bdbdbd", "black", "white")) +
  labs(x = "", y = "Relative abundance")
  supergroup + phylum + patchwork::plot_layout(ncol = 1)
}

# make_bar_relabun(insitu_asv_wClass, axial)
```

## Tile plot

Relative abundance plots are misleading, as this tag-sequence data is
compositional. To combat this, we can also perform a center log-ratio
transformation of the sequence counts. This tile plot (or heat map) will
show the relationship from the data mean. Positive values thus
demonstrate an increase in the taxa, while negative values illustrate
the opposite.

Ahead of the CLR transformation, average across replicates, then sum to
the Class level. THEN perform CLR transformation and plot as heat map.

```{r CLR transformation and generate tile plot}
make_clr_trans_tile <- function(df, selection){
  df_wide <- df %>%
    filter(SITE %in% selection) %>%
  # df_wide <- insitu_asv_wClass %>% 
    # filter(SITE %in% axial) %>% 
    filter(Domain == "Eukaryota") %>% 
  # Average across replicates
      group_by(FeatureID, SAMPLENAME, VENT, Domain, Supergroup, Phylum, Class, Order, Family, Genus, Species) %>% 
        summarise(AVG = mean(value)) %>% 
      ungroup() %>% 
    # Sum to the Order taxonomic classification
    unite(SAMPLENAME_2, SAMPLENAME, VENT, sep = "_") %>% 
      group_by(SAMPLENAME_2, Supergroup, Phylum, Class) %>% 
        summarise(CLASS_SUM = sum(AVG)) %>% 
    unite(CLASS, Supergroup, Phylum, Class, sep = " ") %>% 
    select(CLASS, SAMPLENAME_2, CLASS_SUM) %>% 
    pivot_wider(names_from = SAMPLENAME_2, values_from = CLASS_SUM, values_fill = 0) %>% 
    column_to_rownames(var = "CLASS")
  ## Take wide data frame and CLR transform, pivot to wide, and plot
  data.frame(compositions::clr(df_wide)) %>% 
      rownames_to_column(var = "CLASS") %>%
      pivot_longer(cols = starts_with(selection), values_to = "CLR", names_to = "SAMPLENAME_2") %>% 
      separate(SAMPLENAME_2, c("SAMPLENAME", "VENT"), sep = "_") %>% 
      separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp"), remove = TRUE) %>% 
      mutate(VENT = str_replace_all(VENT, "\\.", " ")) %>% 
      mutate(REGION = case_when(
        SITE == "GordaRidge" ~ "Gorda Ridge",
        SITE %in% mcr ~ "Mid-Cayman Rise",
        # SITE == "Piccard" ~ "Piccard",
        # SITE == "VonDamm" ~ "Von Damm",
        SITE == "Axial" ~ "Axial")) %>% 
      mutate(VENTNAME = case_when(
        REGION == "Gorda Ridge" ~ VENT,
        REGION == "Mid-Cayman Rise" ~ paste(SITE, VENT, sep = " "),
        REGION == "Axial" ~ paste(VENT, YEAR, sep = " ")
      )) %>% select(-Sample_tmp) %>% 
    unite(SAMPLE, REGION, VENTNAME, sep = " ", remove = FALSE) %>% 
    separate(CLASS, c("Supergroup", "Phylum", "Class"), sep = " ", remove = FALSE) %>%
    ggplot(aes(x = SAMPLE, y = Class, fill = CLR)) +
      geom_tile(color = "#252525") + 
      theme(legend.position = "right", 
            panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(), 
            panel.border = element_blank(), 
            panel.background = element_blank(),
            axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, color = "black",size = 8), 
            axis.text.y = element_text(color = "black", size = 8),
            strip.background = element_blank(), 
            strip.text.y = element_text(hjust = 0, vjust = 0.5, angle = 0),
            # strip.text.x = element_blank(), 
            legend.title = element_blank()) + 
      labs(x = "", y = "") +
      scale_fill_gradient2(low = "#4575b4", mid = "white", high = "#d73027", na.value = "grey50") +
      facet_grid(Supergroup + Phylum ~ SAMPLETYPE, space = "free", scales = "free")
}

```

## PCA

Similar to aove, the first step in this function transforms data using
CLR (to ASV level though). First plot will show eigen values (scree plot
to determine if 2 vs. 3 dimensions is best for data). Then function
extracts data points and creates PCA plot.

```{r function to perform CLR transformation and PCoA analysis}
make_pca <- function(df, selection){
 df_wide_asv <- df %>%
    # df_wide_asv <- insitu_asv_wClass %>%
  filter(SITE %in% selection) %>%
    # filter(SITE %in% axial) %>%
  filter(Domain == "Eukaryota") %>% 
  # Average across replicates
    group_by(FeatureID, SAMPLENAME, VENT) %>% 
    summarise(AVG = mean(value)) %>% 
    ungroup() %>% 
  separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp"), remove = TRUE) %>% 
  mutate(REGION = case_when(
    SITE == "GordaRidge" ~ "Gorda Ridge",
    SITE %in% mcr ~ "Mid-Cayman Rise",
    SITE == "Axial" ~ "Axial")) %>% 
  mutate(VENTNAME = case_when(
    REGION == "Gorda Ridge" ~ VENT,
    REGION == "Mid-Cayman Rise" ~ paste(SITE, VENT, sep = " "),
    REGION == "Axial" ~ paste(VENT, YEAR, sep = " ")
  )) %>% select(-Sample_tmp) %>% 
  unite(SAMPLE, SAMPLETYPE, REGION, VENTNAME, sep = "_", remove = FALSE) %>% 
   group_by(FeatureID, SAMPLE) %>% 
   summarise(SUM = sum(AVG)) %>% 
  pivot_wider(names_from = FeatureID, values_from = SUM, values_fill = 0) %>% 
  column_to_rownames(var = "SAMPLE")
  # look at eigenvalues
  pca_lr <- prcomp(data.frame(compositions::clr(df_wide_asv)))
  variance_lr <- (pca_lr$sdev^2)/sum(pca_lr$sdev^2)
  ## View bar plot
  barplot(variance_lr, main = "Log-Ratio PCA Screeplot", xlab = "PC Axis", ylab = "% Variance", 
    cex.names = 1.5, cex.axis = 1.5, cex.lab = 1.5, cex.main = 1.5)
  ## Extract PCR points
  data.frame(pca_lr$x, SAMPLE = rownames(pca_lr$x)) %>% 
      separate(SAMPLE, c("SAMPLETYPE", "REGION", "VENTNAME"), sep = "_", remove = FALSE) %>% 
  ## Generate PCA plot
  ggplot(aes(x = PC1, y = PC2, shape = SAMPLETYPE, fill = VENTNAME)) +
    geom_hline(yintercept = 0) + geom_vline(xintercept = 0, color = "#525252") +
    geom_point(size=4, stroke = 1, aes(fill = VENTNAME)) +
    ylab(paste0('PC2 ',round(variance_lr[2]*100,2),'%')) +
    xlab(paste0('PC1 ',round(variance_lr[1]*100,2),'%')) +
    scale_shape_manual(values = c(21, 23, 24)) +
    scale_fill_viridis(discrete = TRUE, option = "turbo") +
    # scale_fill_manual(values = fill_color) +
    # scale_color_manual(values = color_color) +
    theme_bw() +
    theme(axis.text = element_text(color="black", size=12),
          legend.title = element_blank(),
          axis.title = element_text(color="black", size=14),
          legend.text = element_text(color = "black", size = 14),
          plot.margin = margin(2, 1, 2, 1, "cm")) +
    guides(fill = guide_legend(override.aes = list(shape = 21) ),
              shape = guide_legend(override.aes = list(fill = "black")))
  }
```

## ASV richness

From complete dataset, average across replicates, then sum the total
number of ASVs in each sample. Then plot a data point for total number
of ASVs (ASV richness) by sample type - where sample type represents the
vent, plume, vs. background. Box plots show the median and range.

```{r function to plot boxplot of ASV richness}
make_asv_rich <- function(df, selection){
  df %>% 
  filter(SITE %in% selection) %>%
  filter(Domain == "Eukaryota") %>% 
  # Average across replicates
    group_by(FeatureID, SAMPLENAME, VENT, Supergroup) %>% 
    summarise(AVG = mean(value)) %>% 
  separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp"), remove = TRUE) %>% 
  mutate(REGION = case_when(
    SITE == "GordaRidge" ~ "Gorda Ridge",
    SITE %in% mcr ~ "Mid-Cayman Rise",
    SITE == "Axial" ~ "Axial")) %>% 
  mutate(VENTNAME = case_when(
    REGION == "Gorda Ridge" ~ VENT,
    REGION == "Mid-Cayman Rise" ~ paste(SITE, VENT, sep = " "),
    REGION == "Axial" ~ paste(VENT, YEAR, sep = " ")
  )) %>% select(-Sample_tmp) %>% 
  unite(SAMPLE, REGION, VENTNAME, sep = " ", remove = FALSE) %>% 
  ungroup() %>% 
  group_by(SITE, REGION, SAMPLE, SAMPLETYPE) %>% 
    summarise(NUM_ASV = n()) %>% 
  ggplot(aes(x = SAMPLETYPE, y = NUM_ASV, shape = SAMPLETYPE)) +
  geom_boxplot(aes(group = SAMPLETYPE), alpha = 0.8, width = 0.4) +
  geom_jitter(size=2, aes(fill = SITE)) +
  scale_shape_manual(values = c(21, 23, 24)) +
  theme_bw() +
  theme(axis.text = element_text(color="black", size=12),
        legend.title = element_blank(),
        axis.title = element_text(color="black", size=14),
        legend.text = element_text(color = "black", size = 14)) +
  guides(fill = guide_legend(override.aes = list(shape = 21) ),
            shape = guide_legend(override.aes = list(fill = "black") ) ) +
  labs(x = "", y = "Total number of ASVs")
}
```

## Presence-absence

Bar plot (colors correspond to Supergroup) represents the number of ASVs
shared or unique to each sample. Combination matrix below bars shows
which samples are considered for the bar plot.

```{r function to generate upsetR presence absence plot}
make_upset_plot <- function(df, selection){ 
  df %>% 
  filter(SITE %in% selection) %>%
  filter(Domain == "Eukaryota") %>% 
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  # Taxa to supergroup
  mutate(SupergroupPhylum = SUPERGROUP) %>%
  # Average across replicates
    group_by(FeatureID, SAMPLENAME, VENT, SupergroupPhylum) %>% 
    summarise(AVG = mean(value)) %>% 
  ungroup() %>% 
  separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp"), remove = TRUE) %>% 
  mutate(REGION = case_when(
    SITE == "GordaRidge" ~ "Gorda Ridge",
    SITE %in% mcr ~ "Mid-Cayman Rise",
    SITE == "Axial" ~ "Axial")) %>% 
  mutate(VENTNAME = case_when(
    REGION == "Gorda Ridge" ~ VENT,
    REGION == "Mid-Cayman Rise" ~ paste(SITE, VENT, sep = " "),
    REGION == "Axial" ~ paste(VENT, YEAR, sep = " ")
  )) %>% select(-Sample_tmp) %>% 
  unite(SAMPLE, SITE, SAMPLETYPE, VENT, sep = " ", remove = FALSE) %>% 
  group_by(FeatureID, SupergroupPhylum, SAMPLE) %>% 
    summarise(SUM = sum(AVG)) %>%
  # filter(SUM > 200) %>% 
  ungroup() %>%
  distinct(FeatureID, SupergroupPhylum, SUM, SAMPLE, .keep_all = TRUE) %>% 
  group_by(FeatureID, SupergroupPhylum) %>% 
  summarise(SAMPLE = list(SAMPLE)) %>% 
  ggplot(aes(x = SAMPLE)) +
    geom_bar(color = "black", width = 0.7, aes(fill = SupergroupPhylum)) +
    scale_x_upset(n_intersections = 25) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = "", y = "Shared at ASV level") +
  theme_linedraw() +
  theme(axis.text.y = element_text(color="black", size=14, face = "bold"),
        axis.text.x = element_text(color="black", size=14, face = "bold"),
        axis.title = element_text(color="black", size=14, face = "bold"),
        legend.text = element_text(color = "black", size = 12, face = "bold"),
        legend.title = element_blank(),
        panel.grid.minor = element_blank(),
        plot.margin = margin(1, 1, 1, 5, "cm")) +
  scale_fill_manual(values = all_taxa_color)
}
# all_taxa_color
#values = c("#fa9fb5", "#c51b8a", "#67000d", "#ef3b2c", "#ffffcc", "#feb24c", "#c7e9b4", "#1d91c0", "#deebf7", "#253494", "#9e9ac8", "#238b45", "#54278f", "#bdbdbd", "#252525", "#fa9fb5", "#c51b8a", "#67000d", "#ef3b2c", "#ffffcc", "#feb24c", "#c7e9b4", "#1d91c0", "#253494", "#9e9ac8", "#238b45", "#54278f", "#bdbdbd", "#252525")

```

Filter data to reduce noise and show sample type to vent ecosystem variability.
```{r, fig.height=2, fig.width=8}
alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")
bkgd <- c("Deep seawater", "BSW", "Shallow seawater")
plume <- c("Candelabra Plume", "Mt Edwards Plume", "Plume", "Near vent BW")
cosmo <- c("Vent, plume, & background", "Vent & background", "Vent & plume", "Plume & background")

##
tree_res_v_cosmo <- function(df, site){
  to_supergroup_vent_only <- df %>% 
    filter(SITE %in% site) %>% 
    mutate(DISTRIBUTION = case_when(
    CLASS %in% cosmo ~ "Cosmopolitan",
    CLASS == "Vent only" ~ "Resident"
    )) %>% 
    filter(Domain == "Eukaryota") %>%
    filter(Supergroup != "Opisthokonta") %>% 
    mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
           Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
           Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
           Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
    mutate(SUPERGROUP = case_when(
      Supergroup %in% alv ~ "Other Alveolata",
      Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
      Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
      Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
      Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
      Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
      Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
      TRUE ~ Supergroup
    )) %>% 
    mutate(SAMPLETYPEORDER = case_when(
      VENT %in% bkgd ~ "Background",
      VENT %in% plume ~ "Plume",
      TRUE ~ "Vent"
    )) %>% 
    filter((DISTRIBUTION == "Resident" | DISTRIBUTION == "Cosmopolitan")) %>% 
    group_by(FeatureID, Taxon, SUPERGROUP,
             VENT, SITE, SAMPLETYPE, YEAR, DATASET, SAMPLETYPEORDER, DISTRIBUTION) %>% 
    summarise(SEQ_AVG_REP = mean(value)) %>% 
    ungroup() %>% 
    group_by(SITE, SUPERGROUP, DISTRIBUTION) %>% 
    summarise(ASV_COUNT = n(),
              SEQ_SUM = sum(SEQ_AVG_REP))
  ##
  to_supergroup_vent_only %>% 
    ggplot(aes(area = SEQ_SUM, fill = SUPERGROUP, subgroup = SUPERGROUP)) +
    geom_treemap(color = "white") +
    geom_treemap_subgroup_border(colour = "white", size = 2) +
    # geom_bar(stat = "identity", position = "fill", color = "black", width = 0.9) +
    facet_grid(SITE ~ DISTRIBUTION) +
    theme_linedraw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
          strip.background = element_blank(), strip.text = element_text(color = "black"),
          legend.position = "right",
          legend.title = element_blank(),
          panel.border = element_blank()) +
    scale_y_continuous(expand = c(0,0)) +
    scale_fill_manual(values = c("#fa9fb5", "#c51b8a", "#67000d", "#ef3b2c", "#ffffcc", "#feb24c", "#c7e9b4", "#1d91c0", "#253494", "#9e9ac8", "#238b45", "#54278f", "#bdbdbd", "#252525", "#737373", "#fa9fb5", "#c51b8a", "#67000d", "#ef3b2c", "#ffffcc", "#feb24c", "#c7e9b4", "#1d91c0", "#253494", "#9e9ac8", "#238b45", "#54278f", "#bdbdbd", "#252525")) +
    labs(x = "", y = "Sequence proportion by Supergroup")
}
```

## Top 15 taxa
Isolate the top 15 taxa
```{r, fig.height=8, fig.width=10}
# head(insitu_asv_wClass)

all_class <- as.character(unique(insitu_asv_wClass$CLASS))

all_class_site <- as.character(unique(insitu_asv_wClass$SITE_CLASS))

top15 <- function(df, NUM, site, level, taxa, class, category, plot_tax){
  level <- enquo(level)
  class <- enquo(class)
  plot_tax <- enquo(plot_tax)
  all_class <- as.character(unique(insitu_asv_wClass$CLASS))
  all_class_site <- as.character(unique(insitu_asv_wClass$SITE_CLASS))
out_table <- df %>% 
  # filter(SITE %in% site) %>%
  filter(!!level == taxa) %>%
  # filter(Domain %in% "Eukaryota") %>% 
  # filter(CLASS %in% all_class) %>% 
  filter(!!class %in% category) %>%
  # Average across replicates
  group_by(FeatureID, Taxon, Domain, Supergroup, Phylum, Class, Order, Family, Genus, Species,
           VENT, SITE, SAMPLETYPE, YEAR, DATASET, !!class) %>% 
    summarise(SEQ_AVG_REP = mean(value)) %>% 
  ungroup() %>% 
  # Select top 15 ASVs
  group_by(VENT, SITE, SAMPLETYPE, YEAR, DATASET) %>% 
  top_n(NUM, wt = SEQ_AVG_REP) %>% 
  unite(SAMPLENAME, SITE, SAMPLETYPE, VENT, YEAR, sep = " ", remove = FALSE)
  # 
  plot <- ggplot(out_table, aes(x = SAMPLENAME, fill = !!plot_tax)) +
    geom_bar(stat = "count", color = "black", width = 0.7) +
    coord_flip() +
    facet_grid(SAMPLETYPE + SITE ~ ., space = "free", scales = "free") +
    theme_classic() +
    scale_fill_viridis(discrete = TRUE, option = "turbo")
  ##
  plotly::ggplotly(plot)
}
  
# Function usage:
# top15 <- function(df, NUM, site, level, taxa, class, category, plot_tax)
# top15(insitu_asv_wClass, 10, all, Phylum, "Ciliophora", CLASS, all_class, Class)
# all_class
```


# Axial Seamount analysis

-   What should I do with the year to year comparison?
Individual analyses here for Axial, MCR, and GR will go in supplementary.

## Bar plot relative abundance: Axial

```{r Barplot axial, tidy=FALSE, fig.height=10, fig.width=11}
make_bar_relabun(insitu_asv_wClass, axial)
```

Axial Seamount samples from archived material - span 2013, 2014, and
2015. First, the background and plume (from 2015 only, and from plume
associated with the Anemone vent) are different from the vent samples -
overwhelmingly stramneopile and rhizaria. For the background and plume,
the stramenopiles appear to be associated with ochrophyta or opalozoa.
For the plume, the rhizaria population was associated with cercozoa,
while the background seawater was identified as belonging to radiolaria.

The major difference between the background/plume and vent sites was the
higher relative sequence abundance of ciliates and opisthokonta. For the
opisthokonta, these are primarily metazoa - and I will need to
investigate this further. Exceptions for this include the 'Dependable'
vent from 2013, which had a completely different composition, and
'Marker 113' in 2015, which the opisthokonta sequences were assigned
choanoflagellate and fungi.

*Further questions to consider*

> Any geochemical changes to Marker 113 from 2013/2014 to 2015? Could
 attribute difference of opisthokonta colonization.


## Tile plot CLR: Axial

```{r Isolate axial data and CLR, fig.height=18, fig.width=7}
# make_clr_trans_tile(insitu_asv_wClass, axial)
```

Tile plot goes to the Class taxonomic level. Here at Axial, mostly the
ciliate class had higher CLR values (more enriched relative to the data
mean). Second to ciliates were cercozoa. Also noticing how Marker 113
2013 and 2015 are more similar to each other than 2014?

## PCA: Axial

```{r est eigenvalues axial, fig.height=6, fig.width=8}
make_pca(insitu_asv_wClass, axial)
```

While we only have 1 plume and bsw each for Axial, they are grouping
together - away from vents. So that is an expected signature and likely
consistent with the other sites. These colors are a little confusing, it
does look like Boca is an outlier.

## ASV richness: Axial

```{r Plot ASV richness for axial, tidy=FALSE, fig.height=4, fig.width=6}
# make_asv_rich(insitu_asv_wClass, axial)
```

We only have 1 sample for background and plume from Axial Seamount. But
this shows that the vent sites have varied ASV richness,

## Presence-absence: Axial

```{r generate upsetR plot for axial, fig.width=13, fig.height=9}
make_upset_plot(insitu_asv_wClass, axial)
axial_upset <- make_upset_plot(insitu_asv_wClass, axial)
```

### Presence-absence at Axial - year and site
```{r, fig.width=12, fig.height=9}
alv <- c("Alveolata-Ellobiopsidae", "Alveolata-Perkinsea", "Alveolata-Unknown", "Alveolata-Chrompodellids", "Alveolata-Apicomplexa")
# svg("upsetR-bysite-sampletype-nov2.svg", h=9, w=15)
axial_loc_yr <- insitu_asv_wClass %>% 
  filter(SITE %in% axial) %>%
  filter(Domain == "Eukaryota") %>% 
  filter(Supergroup != "Opisthokonta") %>% 
  mutate(Supergroup = ifelse(is.na(Supergroup), "Unknown Eukaryota", Supergroup),
         Phylum = ifelse(is.na(Phylum), "Unknown", Phylum),
         Phylum = ifelse(Phylum == "Alveolata_X", "Ellobiopsidae", Phylum),
         Supergroup = ifelse(Supergroup == "Alveolata", paste(Supergroup, Phylum, sep = "-"), Supergroup)) %>% 
  mutate(SUPERGROUP = case_when(
    Supergroup %in% alv ~ "Other Alveolata",
    Supergroup == "Eukaryota_X" ~ "Unknown Eukaryota",
    Phylum == "Cercozoa" ~ "Rhizaria-Cercozoa",
    Phylum == "Radiolaria" ~ "Rhizaria-Radiolaria",
    Phylum == "Ochrophyta" ~ "Stramenopiles-Ochrophyta",
    Phylum == "Opalozoa" ~ "Stramenopiles-Opalozoa",
    Phylum == "Sagenista" ~ "Stramenopiles-Sagenista",
    TRUE ~ Supergroup
  )) %>% 
  # Taxa to supergroup
  mutate(SupergroupPhylum = SUPERGROUP) %>%
  # Average across replicates
    group_by(FeatureID, SAMPLENAME, VENT, SupergroupPhylum) %>% 
    summarise(AVG = mean(value)) %>% 
  ungroup() %>% 
  separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp"), remove = TRUE) %>% 
  mutate(REGION = case_when(
    SITE == "GordaRidge" ~ "Gorda Ridge",
    SITE %in% mcr ~ "Mid-Cayman Rise",
    SITE == "Axial" ~ "Axial")) %>% 
  mutate(VENTNAME = case_when(
    REGION == "Gorda Ridge" ~ VENT,
    REGION == "Mid-Cayman Rise" ~ paste(SITE, VENT, sep = " "),
    REGION == "Axial" ~ paste(VENT, YEAR, sep = " ")
  )) %>% select(-Sample_tmp) %>% 
  filter(REGION == "Axial") %>% 
  unite(SAMPLE, SITE, SAMPLETYPE, sep = " ", remove = FALSE) %>% 
  unite(YR_LOC, SAMPLETYPE, YEAR, sep = " ", remove = FALSE) %>% 
  group_by(FeatureID, SupergroupPhylum, YR_LOC) %>% 
    summarise(SUM = sum(AVG)) %>%
  # filter(SUM > 200) %>% 
  ungroup()
##  
axial_loc_yr %>% 
  distinct(FeatureID, SupergroupPhylum, SUM, YR_LOC, .keep_all = TRUE) %>% 
  group_by(FeatureID, SupergroupPhylum) %>% 
  summarise(SAMPLE = list(YR_LOC)) %>% 
  ggplot(aes(x = SAMPLE)) +
    geom_bar(color = "black", width = 0.5, aes(fill = SupergroupPhylum)) +
    scale_x_upset(n_intersections = 25) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = "", y = "Shared ASVs") +
  theme_linedraw() +
  theme(axis.text.y = element_text(color="black", size=14),
        axis.text.x = element_text(color="black", size=14),
        axis.title = element_text(color="black", size=14),
        legend.text = element_text(color = "black", size = 12),
        plot.margin = margin(1, 1, 1, 5, "cm")) +
  scale_fill_manual(values = c("#fa9fb5", "#c51b8a", "#67000d", "#ef3b2c", "#ffffcc", "#feb24c", "#c7e9b4", "#1d91c0", "#deebf7", "#253494", "#9e9ac8", "#238b45", "#54278f", "#bdbdbd", "#252525", "#fa9fb5", "#c51b8a", "#67000d", "#ef3b2c", "#ffffcc", "#feb24c", "#c7e9b4", "#1d91c0", "#253494", "#9e9ac8", "#238b45", "#54278f", "#bdbdbd", "#252525"))
# dev.off()

```


Extract total number of ASVs from Axial data, and those that were found within the vent fluid only.
```{r}
length(unique(axial_loc_yr$FeatureID))
tmp <- axial_loc_yr %>% 
  filter(grepl("Vent", YR_LOC)) %>% 
  pivot_wider(names_from = YR_LOC, values_from = SUM, values_fill = NA) %>% 
  drop_na()
head(tmp)  
length(unique(tmp$FeatureID))
```

```{r}
tmp %>% 
  pivot_longer(starts_with("Vent ")) %>% 
  group_by(name, SupergroupPhylum) %>% 
  summarise(VALUE = sum(value)) %>% 
ggplot(aes(x = name, y = VALUE, fill = SupergroupPhylum)) +
  geom_bar(stat = "identity")
```

## Tree map distribution: Axial
Plot only cosmopolitan vs. resident ASV relative abundances by Supergroup.
```{r, fig.height=4, fig.width=9}
tree_res_v_cosmo(insitu_asv_wClass, axial)
```


## Top 15 taxa: Axial
```{r}
# Function usage:
# top15 <- function(df, NUM, site, level, taxa, class, category, plot_tax)

top15(insitu_asv_wClass, 15, axial, Domain, "Eukaryota", CLASS, all_class, Phylum)
```
```{r}
unique(insitu_asv_wClass$Phylum)
vent_only <- c("Vent only")
# Isolate top 15 vent-only ASVs from axial - at the phylum level
top15(insitu_asv_wClass, 15, axial, Domain, "Eukaryota", CLASS, vent_only, Phylum)

# Top 10 ciliate only taxa, as most were ciliates
top15(insitu_asv_wClass, 10, axial, Phylum, "Ciliophora", CLASS, vent_only, Class)


# Top 10 ciliate only taxa, as most were ciliates
top15(insitu_asv_wClass, 10, axial, Phylum, "Opalozoa", CLASS, vent_only, Class)
```


# Mid-Cayman Rise

## Bar plot relative abundance: MCR

```{r Barplot MCR, fig.height=10, fig.width=11}
# make_bar_relabun(insitu_asv_wClass, mcr)
```

## Tile plot CLR: MCR

```{r Isolate MCR data and CLR, fig.height=13, fig.width=8}
# make_clr_trans_tile(insitu_asv_wClass, mcr)
```

## PCA: MCR

```{r est eigenvalues MCR, fig.height=6, fig.width=8}
make_pca(insitu_asv_wClass, mcr)
```

## ASV richness: MCR

```{r Plot ASV richness for MCR, tidy=FALSE}
# make_asv_rich(insitu_asv_wClass, mcr)
```

## Presence-absence: MCR

```{r make upsetR plot for MCR, fig.width=13, fig.height=9}
make_upset_plot(insitu_asv_wClass, mcr)
mcr_upset <- make_upset_plot(insitu_asv_wClass, mcr)
```

Repeat presence-absence plot, but with a lower resolution.

```{r, fig.width=8, fig.height=4}
insitu_asv_wClass %>% 
  filter(SITE %in% mcr) %>%
  filter(Domain == "Eukaryota") %>% 
  # Average across replicates
    group_by(FeatureID, SAMPLENAME, VENT, Supergroup) %>% 
    summarise(AVG = mean(value)) %>% 
  separate(SAMPLENAME, c("SITE", "SAMPLETYPE", "YEAR", "Sample_tmp"), remove = TRUE) %>% 
  mutate(REGION = case_when(
    SITE == "GordaRidge" ~ "Gorda Ridge",
    SITE %in% mcr ~ "Mid-Cayman Rise",
    SITE == "Axial" ~ "Axial")) %>% 
  mutate(VENTNAME = case_when(
    REGION == "Gorda Ridge" ~ VENT,
    REGION == "Mid-Cayman Rise" ~ paste(SITE, VENT, sep = " "),
    REGION == "Axial" ~ paste(VENT, YEAR, sep = " ")
  )) %>% select(-Sample_tmp) %>% 
  unite(SAMPLE, SITE, SAMPLETYPE, sep = " ", remove = FALSE) %>% 
  group_by(FeatureID, Supergroup, SAMPLE) %>% 
    summarise(SUM = sum(AVG)) %>%
  ungroup() %>% 
  distinct(FeatureID, Supergroup, SUM, SAMPLE, .keep_all = TRUE) %>% 
  group_by(FeatureID, Supergroup) %>% 
  summarise(SAMPLE = list(SAMPLE)) %>% 
  ggplot(aes(x = SAMPLE)) +
    geom_bar(color = "black", width = 0.5, aes(fill = Supergroup)) +
    scale_x_upset(n_intersections = 15) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = "", y = "Shared ASVs") +
  theme_linedraw() +
  theme(axis.text = element_text(color="black", size=10),
        axis.title = element_text(color="black", size=10),
        legend.text = element_text(color = "black", size = 10),
        plot.margin = margin(1, 1, 1, 5, "cm")) +
    scale_fill_manual(values = c("#f1eef6", "#d7b5d8", "#df65b0", "#ce1256", "#fc9272", "#ef3b2c", 
    "#800026", "#fff7bc", "#fec44f", "#d95f0e", "#a63603", "#74c476", "#238b45", 
    "#00441b", "#7fcdbb", "#084081", "#c6dbef", "#2b8cbe", "#016c59", "#bcbddc", 
    "#807dba", "#54278f", "#bdbdbd", "black"))
```

## Top 15 taxa: MCR
```{r}
# Function usage:
# top15 <- function(df, NUM, site, level, taxa, class, category, plot_tax)
top15(insitu_asv_wClass, 15, mcr, Domain, "Eukaryota", CLASS, all_class, Phylum)
```

```{r}
# unique(insitu_asv_wClass$CLASS)
vent_only <- c("Vent only")
# Isolate top 15 vent-only ASVs from MCR - at the phylum level
top15(insitu_asv_wClass, 15, mcr, Domain, "Eukaryota", CLASS, vent_only, Phylum)

# Top 10 ciliate only taxa, as most were ciliates
top15(insitu_asv_wClass, 10, mcr, Phylum, "Ciliophora", CLASS, vent_only, Class)
```
# Gorda Ridge

 ## Bar plot relative abundance: GR

```{r Barplot GR, tidy=FALSE, fig.height=10, fig.width=11}
make_bar_relabun(insitu_asv_wClass, gr)
```

## Tile plot CLR: GR

```{r Isolate GR data and CLR, fig.height=12, fig.width=6}
# make_clr_trans_tile(insitu_asv_wClass, gr)
```

## PCA: GR

```{r est eigenvalues GR, fig.height=6, fig.width=8}
make_pca(insitu_asv_wClass, gr)
```

## ASV richness: GR

```{r Plot ASV richness for GR, tidy=FALSE}
# make_asv_rich(insitu_asv_wClass, gr)
```

## Presence-absence: GR

```{r, fig.width=13, fig.height=9}
make_upset_plot(insitu_asv_wClass, gr)
gr_upset <- make_upset_plot(insitu_asv_wClass, gr)
```

## Compile upsetR Fig S3.
```{r, fig.height=17, fig.width=20}
# svg("../../../Manuscripts_presentations_reviews/deepsea-microeuk-ME/vectorgraphic-figures/figs3-upsetRpanels.svg", h = 17, w = 20)
(genus_upset + theme(legend.position = "none")) + (axial_upset + theme(legend.position = "none")) + (gr_upset + theme(legend.position = "none")) + (mcr_upset + theme(legend.position = "none")) + patchwork::plot_layout(nrow = 2, ncol = 2) + plot_annotation(tag_levels = "a")
# dev.off()
```


## Top 15 taxa: GR
```{r}
# Function usage:
# top15 <- function(df, NUM, site, level, taxa, class, category, plot_tax)
top15(insitu_asv_wClass, 15, gr, Domain, "Eukaryota", CLASS, all_class, Phylum)
```

```{r}
# unique(insitu_asv_wClass$CLASS)
vent_only <- c("Vent only")
# Isolate top 15 vent-only ASVs from GR - at the phylum level
top15(insitu_asv_wClass, 15, gr, Domain, "Eukaryota", CLASS, vent_only, Phylum)

# Top 10 ciliate only taxa, as most were ciliates
top15(insitu_asv_wClass, 10, gr, Phylum, "Ciliophora", CLASS, vent_only, Class)
```

# Session Info
```{r}
sessionInfo()
```