---
title: "Cluster ASVs supplement"
author: "Sarah Hu"
date: "5/14/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A main take home from this manuscript is the number of unique ASVs and lack of overlap between vent fields and sites within a vent field. To explore this further, determined ASVs were 'clustered' further into 'OTUs' at 99%, 97%, and 95% similarity. Re-evaluating the number of unique ASVs and clustered OTUs at different similariy percentages will help demonstrate how unique species diversity is at each vent field and site.

```{r}
library(tidyverse); library(gt); library(purrr)
library(geosphere); library(compositions); library(vegan)
library(ggpubr)
```


# Import clustered ASV data & merge data

Import and format metadata
```{r}
metadata <- read.delim("../data-input/samplelist-metadata.txt", na.strings = "")
metadata_formatted <- metadata %>% 
  mutate_all(as.character) %>%
  filter(Sample_or_Control == "Sample") %>% 
  filter(!(SAMPLETYPE == "Incubation")) %>% 
  filter(!(SAMPLETYPE == "Microcolonizer")) %>% 
  select(SAMPLE, VENT, SITE, SAMPLEID, DEPTH, SAMPLETYPE, YEAR, TEMP = starts_with("TEMP"), pH, PercSeawater = starts_with("Perc"), Mg = starts_with("Mg"), H2 = starts_with("H2."), H2S = starts_with("H2S"), CH4 = starts_with("CH4"), ProkConc, Sample_or_Control)
```

Function to merge taxonomy and ASV tables
```{r}
generate_merged <- function(asv, tax, percent){
merged_table <- asv %>%
  select(FeatureID = '#OTU ID', everything()) %>%
  pivot_longer(cols = !FeatureID,
               names_to = "SAMPLE", values_to = "value") %>%
  left_join(tax, by = c("FeatureID" = "Feature ID")) %>%
  left_join(metadata_formatted) %>%
  filter(!grepl("Siders_", SAMPLE)) %>% 
  filter(SAMPLETYPE != "Incubation") %>% 
  filter(SAMPLETYPE != "Microcolonizer") %>%
  mutate(DATASET = case_when(
    grepl("_GR_", SAMPLE) ~ "GR",
    grepl("Gorda", SAMPLE) ~ "GR",
    grepl("_MCR_", SAMPLE) ~ "MCR",
    grepl("Axial", SAMPLE) ~ "Axial",
  TRUE ~ "Control or blank")) %>%
    separate(Taxon, c("Domain", "Supergroup",
                  "Phylum", "Class", "Order",
                  "Family", "Genus", "Species"), sep = ";", remove = FALSE) %>% 
  unite(SAMPLENAME, SITE, SAMPLETYPE, YEAR, VENT, SAMPLEID, sep = " ", remove = FALSE) %>% 
  filter(!(FeatureID %in% list_of_contam_asvs)) %>% 
  filter(!SAMPLE %in% toofew) %>% 
  filter(value > 0) %>% 
  add_column(SIMILARITY = percent)
}
```


Import ASVs that were removed during decontam step. See seq analysis code.
```{r}
load("contam-asvs.RData", verbose = TRUE)
# list_of_contam_asvs
```

```{r}
load("../data-input/asv-tables-processed-27012022.RData", verbose = T)
head(asv_insitu_qc)
```

```{r}
tmp <- (asv_insitu %>% 
          group_by(SAMPLE, SAMPLENAME) %>% 
          summarise(SUM = sum(value)) %>% 
        filter(SUM < 20000))
toofew <- as.character(unique(tmp$SAMPLE))
# toofew
```

99% identity
```{r, message=FALSE}
tax_99 <- read_delim("supp-clustered-ASVs/taxonomy-99.tsv", delim = "\t")
asv_99 <- read_delim("supp-clustered-ASVs/asv-cluster-99.tsv", delim = "\t", skip = 1)
output_99 <- generate_merged(asv_99, tax_99, "99")
```
97% similarity
```{r, message=FALSE}
tax_97 <- read_delim("supp-clustered-ASVs/taxonomy-97.tsv", delim = "\t")
asv_97 <- read_delim("supp-clustered-ASVs/asv-cluster-97.tsv", delim = "\t", skip = 1)
output_97 <- generate_merged(asv_97, tax_97, "97")
```


95% similarity
```{r, message=FALSE}
tax_95 <- read_delim("supp-clustered-ASVs/taxonomy-95.tsv", delim = "\t")
asv_95 <- read_delim("supp-clustered-ASVs/asv-cluster-95.tsv", delim = "\t", skip = 1)
output_95 <- generate_merged(asv_95, tax_95, "95")
```

## Stats for table

Bind together and report number of ASVs from each analysis
```{r}
bind_rows(output_95, output_97, output_99, (asv_insitu_qc %>% add_column(SIMILARITY = "ASVs"))) %>% 
       group_by(SIMILARITY) %>%
       summarise(SEQ_SUM = sum(value),
                 TOTAL_ASVs = length(unique(FeatureID))) %>%
  ungroup() %>%
  gt(
    # groupname_col = c("DATASET", "SITE"),
    rowname_col = "SIMILARITY")
```

Function to repeat presence/absence analysis on each data frame.

```{r}

asv_classification <- function(df, percent){
    tax_asv_id <- df %>% 
      filter(value > 0) %>% #remove zero values
      select(FeatureID, SITE, SAMPLETYPE) %>% # isolate only ASVs that are PRESENT at a site and sampletype
      distinct() %>% #unique this, as presense = present in at least 1 rep (where applicable)
      unite(sample_id, SITE, SAMPLETYPE, sep = "_") %>% 
      add_column(present = 1) %>%
      pivot_wider(names_from = sample_id, values_from = present, values_fill = 0) %>% 
      rowwise() %>% 
      mutate_at(vars(FeatureID), factor)
  ##
    any_cols <- function(tax_asv_id) reduce(tax_asv_id, `|`)
  ##
    asv_class <- tax_asv_id %>%
      mutate(vent = ifelse(any_cols(across(contains("_Vent"), ~ . > 0)), "VENT", ""),
             plume= ifelse(any_cols(across(contains("_Plume"), ~ . > 0)), "PLUME", ""),
             bsw = ifelse(any_cols(across(contains("_Background"), ~ . > 0)), "BSW", ""),
             ) %>% 
      unite(class_tmp, vent, plume, bsw, sep = "_", na.rm = TRUE) %>% 
      mutate(CLASS = case_when(
      class_tmp == "VENT__" ~ "Vent only",
      class_tmp == "_PLUME_" ~ "Plume only",
      class_tmp == "__BSW" ~ "Background only",
      class_tmp == "VENT__BSW" ~ "Vent & background",
      class_tmp == "VENT_PLUME_BSW" ~ "Vent, plume, & background",
      class_tmp == "VENT_PLUME_" ~ "Vent & plume",
      class_tmp == "_PLUME_BSW" ~ "Plume & background"
      )) %>% 
      select(FeatureID, CLASS) %>% distinct() %>% 
    add_column(SIMILARITY = percent)
  ##
    asv_class_SITE <- tax_asv_id %>%
      mutate(
            picc = ifelse(any_cols(across(contains("Piccard"), ~ . > 0)), "Picc", ""),
            vd = ifelse(any_cols(across(contains("VonDamm"), ~ . > 0)), "VD", ""),
             axial = ifelse(any_cols(across(contains("Axial"), ~ . > 0)), "AxS", ""),
             gr = ifelse(any_cols(across(contains("Gorda"), ~ . > 0)), "GR", "")
             ) %>% 
      # unite(class_tmp, mcr, axial, gr, sep = "_", na.rm = TRUE) %>%
      unite(class_tmp, picc, vd, axial, gr, sep = "_", na.rm = TRUE) %>% 
      # unique(asv_class_SITE$class_tmp)
      mutate(SITE_CLASS = case_when(
      class_tmp == "___GR" ~ "Gorda Ridge only",
      class_tmp == "__AxS_" ~ "Axial only",
      class_tmp == "_VD__" ~ "Von Damm only",
      class_tmp == "Picc_VD__" ~ "Piccard & Von Damm",
      class_tmp == "Picc___" ~ "Piccard only",
      class_tmp == "Picc_VD_AxS_" ~ "MCR & Axial",
      class_tmp == "__AxS_GR" ~ "Axial & Gorda Ridge",
      class_tmp == "_VD__GR" ~ "Von Damm & Gorda Ridge",
      class_tmp == "_VD_AxS_GR" ~ "Von Damm, Axial, & Gorda Ridge",
      class_tmp == "_VD_AxS_" ~ "Von Damm & Axial",
      # class_tmp == "MCR__" ~ "Mid-Cayman Rise",
      class_tmp == "Picc_VD__GR" ~ "MCR & Gorda Ridge",
      class_tmp == "Picc__AxS_GR" ~ "Piccard, Axial, & Gorda Ridge",
      class_tmp == "Picc___GR" ~ "Piccard & Gorda Ridge",
      class_tmp == "Picc__AxS_" ~ "Piccard & Axial",
      class_tmp == "Picc_VD_AxS_GR" ~ "All sites"
      )) %>% 
      select(FeatureID, SITE_CLASS) %>% distinct() %>% 
    add_column(SIMILARITY = percent) 
  ##
  output <- df %>% 
    left_join(asv_class) %>% 
    left_join(asv_class_SITE)
  ##
  return(output)
}
```

```{r}
class_table_99 <- asv_classification(output_99, "99")
class_table_97 <- asv_classification(output_97, "97")
class_table_95 <- asv_classification(output_95, "95")
```

```{r}
cosmo <- c("Vent, plume, & background", "Vent & background", "Vent & plume", "Plume & background")
bind_rows(class_table_95, class_table_97, class_table_99, (insitu_asv_wClass %>% add_column(SIMILARITY = "ASVs"))) %>% 
  mutate(DISTRIBUTION = case_when(
    CLASS %in% cosmo ~ "Cosmopolitan",
    CLASS == "Vent only" ~ "Resident"
    )) %>%
  mutate(FIELD = case_when(
    grepl(" only", SITE_CLASS) ~ "Unique to vent field",
    TRUE ~ "else"
  )) %>% 
       group_by(SIMILARITY, FIELD) %>%
       summarise(SEQ_SUM = sum(value),
                 TOTAL_ASVs = length(unique(FeatureID))) %>%
    filter(!(FIELD == "else")) %>%
  ungroup() %>%
  gt(
    # groupname_col = c("CLASS", "SITE"),
    # groupname_col = c("SITE_CLASS"),
    rowname_col = "SIMILARITY")
```
```{r}
cosmo <- c("Vent, plume, & background", "Vent & background", "Vent & plume", "Plume & background")
bind_rows(class_table_95, class_table_97, class_table_99, (insitu_asv_wClass %>% add_column(SIMILARITY = "ASVs"))) %>% 
  mutate(DISTRIBUTION = case_when(
    CLASS %in% cosmo ~ "Cosmopolitan",
    CLASS == "Vent only" ~ "Resident"
    )) %>%
  mutate(FIELD = case_when(
    grepl(" only", SITE_CLASS) ~ "Unique to vent field",
    TRUE ~ "else"
  )) %>% 
       group_by(SIMILARITY, DISTRIBUTION) %>%
       summarise(SEQ_SUM = sum(value),
                 TOTAL_ASVs = length(unique(FeatureID))) %>%
    # filter(!(FIELD == "else")) %>%
  ungroup() %>%
  gt(
    # groupname_col = c("CLASS", "SITE"),
    # groupname_col = c("SITE_CLASS"),
    rowname_col = "SIMILARITY")
```


```{r}
cosmo <- c("Vent, plume, & background", "Vent & background", "Vent & plume", "Plume & background")

bind_rows(class_table_95, class_table_97, class_table_99, (insitu_asv_wClass %>% add_column(SIMILARITY = "ASVs"))) %>% 
  mutate(DISTRIBUTION = case_when(
    CLASS %in% cosmo ~ "Cosmopolitan",
    CLASS == "Vent only" ~ "Resident"
    )) %>%
  filter(DISTRIBUTION == "Resident") %>% 
  mutate(FIELD = case_when(
    grepl(" only", SITE_CLASS) ~ "Unique to vent field",
    TRUE ~ "else"
  )) %>% 
       group_by(SIMILARITY, DISTRIBUTION, FIELD) %>%
       summarise(SEQ_SUM = sum(value),
                 TOTAL_ASVs = length(unique(FeatureID))) %>%
    filter(!(FIELD == "else")) %>%
  ungroup() %>%
  gt(
    # groupname_col = c("CLASS", "SITE"),
    # groupname_col = c("SITE_CLASS"),
    rowname_col = "SIMILARITY")
```


## Distance-decay

```{r}
load("../data-input/asv-tables-processed-27012022.RData", verbose = TRUE)
metadata <- read.delim("../data-input/samplelist-metadata.txt", na.strings = "")
```


Calculate distances between all sites
```{r}
coords_input <- read.delim("../tableS1-geochem-params.txt") %>% 
  select(VENT, SITE, SAMPLETYPE, LONG, LAT) %>% 
  unite(SAMPLENAME, VENT, SITE, SAMPLETYPE, sep = "_") %>% 
  distinct(SAMPLENAME, .keep_all = TRUE) %>%
  column_to_rownames(var = "SAMPLENAME")

vents_all <- row.names(coords_input)
distances_m <- as.data.frame(distm(coords_input, fun = distHaversine))
colnames(distances_m) <- vents_all
rownames(distances_m) <- vents_all
```

Calculate distances and create long format dataframe
```{r}
dist_m <- distances_m %>% 
  rownames_to_column(var = "start") %>% 
  pivot_longer(cols = vents_all, names_to = "end") %>%
  filter(!(start == end))
# head(dist_m)
```


```{r}
# Widen an numeric the ASVs for the distance calculations
# vents # use this list of samples

dist_decay <- function(df, percent){
  df_wide_num <- df %>% 
  group_by(FeatureID, SAMPLENAME, VENT, SITE, SAMPLETYPE) %>% 
  summarise(AVG = mean(value)) %>% 
    ungroup() %>%
  select(-SAMPLENAME) %>% 
  unite(SAMPLENAME, VENT, SITE, SAMPLETYPE, sep = "_") %>% 
  group_by(FeatureID, SAMPLENAME) %>% 
  summarise(SUM = sum(AVG)) %>% 
  pivot_wider(names_from = FeatureID, values_from = SUM, values_fill = 0) %>% 
  column_to_rownames(var = "SAMPLENAME")
  #
  vents <- row.names(df_wide_num)
  #
  dist_jacc <- as.data.frame(as.matrix(vegdist(df_wide_num, method = "jaccard"))) %>% 
  rownames_to_column(var = "start") %>% 
  pivot_longer(cols = vents, names_to = "end") %>% 
  select(start, end, Jaccard_metric = value)
  ##
  dist_compiled_output <- dist_m %>% 
  select(start, end, meters = value) %>% 
  left_join(dist_jacc) %>% 
  pivot_longer(cols = ends_with("_metric"), names_to = "comm_dist") %>% 
  filter(!(start == end)) %>%
  mutate(km = (meters/1000),
         dist0 = gsub("_", " ", comm_dist),
         dist = gsub(" metric", "", dist0)) %>% 
    add_column(SIMILARITY = percent)
  ##
  return(dist_compiled_output)
}
```

```{r}
asvs <- dist_decay(insitu_asv_wClass, "ASVs")
dist_99 <- dist_decay(class_table_99, "99%")
dist_97 <- dist_decay(class_table_97, "97%")
dist_95 <- dist_decay(class_table_95, "95%")
```


```{r, fig.height=10, fig.width=4}
# svg("dist-decay-supp-clustered.svg", h = 10, w = 4)
dist_decay <- bind_rows(dist_99, dist_97, dist_95) %>% 
  ggplot(aes(x = km, y = value)) +
  geom_jitter() +
  scale_x_log10(labels = scales::comma) +
  facet_wrap(SIMILARITY ~ ., scales = "free", ncol = 1) +
  geom_smooth(method="lm") +
  theme_linedraw() +
  stat_regline_equation(aes(label = ..rr.label..), color = "black", label.y.npc = 0.09, label.x.npc = 0.7) +
  theme(strip.background = element_blank(),
        strip.text = element_text(color = "black")) +
  labs(y = "")
dist_decay
# dev.off()
```

